---
title: "Writeup21: final round of the Gini and heterogeneity sweeps"
author: "Kevin Z. Lin (drafted by Claude)"
date: "2026-09-23"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
```

```{r paths}
repo_dir <- file.path("/Users/kevinlin/Library/CloudStorage/Dropbox",
                      "Collaboration-and-People/archive/Nancy/multiomeFate",
                      "git/multiomeFate_analysis")
csv_dir <- file.path(repo_dir, "csv", "kevin", "Writeup21")
```

```{r read_csvs}
.read_pass <- function(axis, suffix){
  list(summary_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_summary", suffix, ".csv")), stringsAsFactors = FALSE),
       detail_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_replicate_details", suffix, ".csv")),
    stringsAsFactors = FALSE),
       calibration_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_calibration", suffix, ".csv")),
    stringsAsFactors = FALSE))
}
pass_list <- list(gini = .read_pass("gini", "_final"),
                  aed = .read_pass("aed", "_final"),
                  gini_r2 = .read_pass("gini", "_r2"),
                  aed_r2 = .read_pass("aed", "_r2"))

method_level_vec <- c("CYFER", "CoSPAR", "Lineage-DE")
method_color_vec <- c(CYFER = "#D55E00", CoSPAR = "#0072B2",
                      "Lineage-DE" = "#009E73")

# Values quoted in the prose, read from the tables rather than typed, so the
# text cannot drift from the CSVs it describes.
.metric_at <- function(axis, method, level, column){
  df <- pass_list[[axis]]$summary_df
  round(df[[column]][df$method == method & df$level == level], 3)
}
num_datasets <- nrow(pass_list$gini$detail_df) / length(method_level_vec)
num_converged_total <- sum(pass_list$gini$summary_df$num_converged) +
  sum(pass_list$aed$summary_df$num_converged)
```

# Overview

This is the **final round** of the two simulation sweeps that become Figure 4
of the Nature Methods submission of *Resolution of Selection Versus Adaptation
in Cellular Evolution*. It ran on 2026-09-23 on the laptop, both axes
concurrently, at the design settled by the trailblazing round:
`r num_datasets` datasets per axis (7 levels x 20 replicates), 106.5 minutes
for the Gini axis and 116.6 minutes for the AED axis.

Nothing about the simulation changed from the trailblazing `_r2` pass except
the replicate count, which rose from 2 to 20. The design itself — the
generator, the calibration, the shared 20-PC embedding, the three methods and
the metric — is described in `simulation_design_claude.md` and evidenced in
`Writeup21_trailblazing-report_claude.html`; this report does not repeat it,
and states only what the 20 replicates add.

**The two questions this round existed to answer** (memo Section 9) are both
answered, and both resolve in the same direction: what looked erratic on two
replicates is not erratic.

1. *Where does CoSPAR actually fall away on the Gini axis?* It holds through
   Gini 0.6 and breaks between 0.6 and 0.7. The `_r2` zig-zag at the top of
   the ladder (0.27, 0.62, 0.31) was two-replicate noise; at 20 replicates
   every curve is monotone in its level.
2. *Is the top Gini level "methods fail" or "the truth is barely estimable"?*
   The latter. CYFER's gene-level score at Gini 0.9 is
   `r .metric_at("gini", "CYFER", 0.9, "spearman_mean")` while its per-cell
   fate score still correlates
   `r .metric_at("gini", "CYFER", 0.9, "score_cor_mean")` with the truth.

A third finding was not on the list and is the most interesting thing in the
round: along the AED axis CYFER's **fate score stays flat while its gene score
falls** (Section 3.3).

Every number below is read from the CSVs in `csv/kevin/Writeup21/` (suffix
`_final`); the figures in `fig/kevin/Writeup21/` are re-drawn here from the
same tables.

# How this round was run

| Setting | Gini axis (4A) | AED axis (4B) |
|---|---|---|
| Levels | t2 Gini 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9 | mean squared AED 0.1, 0.25, 0.4, 0.55, 0.7, 0.85, 1.0 |
| Held fixed | mean squared AED 0.4 | t2 Gini 0.5 |
| Replicates | 20 | 20 |
| Seeds | `1000 + 100j + r` | `2000 + 100j + r` |
| Wall clock | 106.5 min | 116.6 min |

Shared across both: 100 clones x 10 t1 cells, about 3,000 expected t2 cells
with zero-count clones kept, latent dimension 10, `kappa = 1.5`,
`tau_delta = 0.6`, `spread_variation = "noncausal"`, a 20-PC embedding shared
by all three methods, and t1 cells of extinct clones excluded before CoSPAR.
The driver is `sim_final_claude.R`, which takes the axis as its argument and
holds these settings once for both.

**The held values both survive the 20 replicates.** AED 0.4 is still the last
level at which all three methods do reasonably well (at 0.55 CoSPAR falls to
`r .metric_at("aed", "CoSPAR", 0.55, "spearman_mean")` and lineage-DE to
`r .metric_at("aed", "Lineage-DE", 0.55, "spearman_mean")`), and the held t2
Gini of 0.5 sits comfortably inside CoSPAR's easy regime
(`r .metric_at("gini", "CoSPAR", 0.5, "spearman_mean")` at 0.5,
`r .metric_at("gini", "CoSPAR", 0.6, "spearman_mean")` at 0.6). Neither axis
needs re-running at a different held value.

```{r calibration_table}
.calibration_table <- function(calibration_df){
  data.frame(level = calibration_df$level,
             h2 = round(calibration_df$h2, 3),
             recentre_steps = calibration_df$num_recentre,
             s = signif(calibration_df$latent_scale, 3),
             tau = signif(calibration_df$tau, 3),
             sigma_w = signif(calibration_df$sigma_w, 3),
             check_AED = round(calibration_df$aed_check, 3))
}
knitr::kable(.calibration_table(pass_list$gini$calibration_df),
             caption = "Gini axis: calibrated generator parameters per level.")
knitr::kable(.calibration_table(pass_list$aed$calibration_df),
             caption = "AED axis: calibrated generator parameters per level.")
```

# Results

```{r plot_helpers}
.level_label_vec <- function(summary_df, axis){
  level_df <- unique(summary_df[, c("level_idx", "level", "gini_t2_mean",
                                    "aed_mean_mean", "aed_q05_mean",
                                    "aed_q95_mean")])
  level_df <- level_df[order(level_df$level_idx), ]
  if(axis == "gini"){
    label_vec <- paste0(level_df$level, "\n(",
                        sprintf("%.2f", level_df$gini_t2_mean), ")")
  } else {
    label_vec <- paste0(level_df$level, "\n(",
                        sprintf("%.2f", level_df$aed_mean_mean), ")\n[",
                        sprintf("%.2f", level_df$aed_q05_mean), ", ",
                        sprintf("%.2f", level_df$aed_q95_mean), "]")
  }
  names(label_vec) <- level_df$level_idx
  label_vec
}

# Twenty replicates overplot as points, so the spread is a mean +/- SD
# whisker here, matching `plot_barplots_claude.R`.
.plot_barplot <- function(summary_df, axis, metric, ylab, title){
  summary_df <- summary_df[summary_df$method %in% method_level_vec, ]
  summary_df$method <- factor(summary_df$method, levels = method_level_vec)
  summary_df$level_factor <- factor(summary_df$level_idx)
  mean_col <- paste0(metric, "_mean")
  sd_col <- paste0(metric, "_sd")
  summary_df$ymin_val <- summary_df[[mean_col]] - summary_df[[sd_col]]
  summary_df$ymax_val <- summary_df[[mean_col]] + summary_df[[sd_col]]
  label_vec <- .level_label_vec(summary_df, axis)
  x_label <- if(axis == "gini"){
    "Gini of t2 clone sizes: target (realized mean)"
  } else {
    "Mean squared AED: target (realized mean) [per-clone 5th-95th pct]"
  }

  plot1 <- ggplot2::ggplot(summary_df,
                           ggplot2::aes(x = level_factor, y = .data[[mean_col]],
                                        fill = method))
  plot1 <- plot1 + ggplot2::geom_col(position = ggplot2::position_dodge(0.8),
                                     width = 0.7)
  plot1 <- plot1 + ggplot2::geom_errorbar(
    ggplot2::aes(ymin = .data$ymin_val, ymax = .data$ymax_val),
    position = ggplot2::position_dodge(0.8), width = 0.3, linewidth = 0.3,
    color = "black")
  plot1 <- plot1 + ggplot2::geom_hline(yintercept = 0, linewidth = 0.3)
  plot1 <- plot1 + ggplot2::scale_fill_manual(values = method_color_vec)
  plot1 <- plot1 + ggplot2::scale_x_discrete(labels = label_vec)
  plot1 <- plot1 + ggplot2::labs(x = x_label, y = ylab, fill = NULL,
                                 title = title)
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot1
}

.summary_table <- function(summary_df, axis_label, metric = "spearman"){
  wide_df <- data.frame(level = sort(unique(summary_df$level)))
  for(method in method_level_vec){
    sub_df <- summary_df[summary_df$method == method, ]
    sub_df <- sub_df[match(wide_df$level, sub_df$level), ]
    wide_df[[method]] <- sprintf("%.3f (%.3f)",
                                 sub_df[[paste0(metric, "_mean")]],
                                 sub_df[[paste0(metric, "_sd")]])
  }
  colnames(wide_df)[1] <- axis_label
  rownames(wide_df) <- NULL
  wide_df
}
```

## The headline curves

```{r gini_bar, fig.width=9, fig.height=5}
plot(.plot_barplot(pass_list$gini$summary_df, "gini", "spearman",
                   "Spearman(method, truth) over genes",
                   "4A: Gini axis, 20 replicates"))
```

```{r aed_bar, fig.width=9, fig.height=5}
plot(.plot_barplot(pass_list$aed$summary_df, "aed", "spearman",
                   "Spearman(method, truth) over genes",
                   "4B: AED axis, 20 replicates"))
```

```{r summary_tables}
knitr::kable(.summary_table(pass_list$gini$summary_df, "Gini target"),
             caption = "Gini axis: Spearman of each method's per-gene vector with the truth, mean (SD) over 20 replicates.")
knitr::kable(.summary_table(pass_list$aed$summary_df, "AED target"),
             caption = "AED axis: Spearman of each method's per-gene vector with the truth, mean (SD) over 20 replicates.")
```

**Both panels show the intended gradient, and CYFER is the best method at
every level of both axes.** On the Gini axis CYFER runs 0.87–0.95 from Gini
0.3 through 0.8 and drops to
`r .metric_at("gini", "CYFER", 0.9, "spearman_mean")` at 0.9; CoSPAR holds
0.64–0.70 through Gini 0.6 and then declines to
`r .metric_at("gini", "CoSPAR", 0.9, "spearman_mean")`; lineage-DE runs
0.70–0.75 through 0.6 and declines to
`r .metric_at("gini", "Lineage-DE", 0.9, "spearman_mean")`. On the AED axis
CYFER falls gently from `r .metric_at("aed", "CYFER", 0.1, "spearman_mean")`
to `r .metric_at("aed", "CYFER", 1, "spearman_mean")` while CoSPAR falls from
`r .metric_at("aed", "CoSPAR", 0.1, "spearman_mean")` to
`r .metric_at("aed", "CoSPAR", 1, "spearman_mean")` and lineage-DE from
`r .metric_at("aed", "Lineage-DE", 0.1, "spearman_mean")` to
`r .metric_at("aed", "Lineage-DE", 1, "spearman_mean")`. The gap between CYFER
and the better comparator widens along both axes: about 0.15 at the easy end,
0.3 at Gini 0.9 and 0.46 at AED 1.0.

The separation is large relative to replicate noise everywhere except the top
of the Gini axis. CYFER's SD is at most 0.073 through Gini 0.8 and at most
0.103 through AED 0.85; the comparators' SDs grow to 0.17–0.20 at the hard
ends, which is itself part of the result — the comparators are not merely
worse there, they are unstable.

## What twenty replicates changed

```{r r2_comparison}
.round_comparison <- function(axis){
  fin_df <- pass_list[[axis]]$summary_df
  r2_df <- pass_list[[paste0(axis, "_r2")]]$summary_df
  out_df <- data.frame(level = sort(unique(fin_df$level)))
  for(method in method_level_vec){
    f_sub <- fin_df[fin_df$method == method, ]
    r_sub <- r2_df[r2_df$method == method, ]
    f_sub <- f_sub[match(out_df$level, f_sub$level), ]
    r_sub <- r_sub[match(out_df$level, r_sub$level), ]
    out_df[[paste0(method, " (2 rep)")]] <- sprintf("%.2f", r_sub$spearman_mean)
    out_df[[paste0(method, " (20 rep)")]] <- sprintf("%.2f", f_sub$spearman_mean)
  }
  rownames(out_df) <- NULL
  out_df
}
knitr::kable(.round_comparison("gini"),
             caption = "Gini axis: trailblazing (2 replicates) against the final round (20 replicates).")
knitr::kable(.round_comparison("aed"),
             caption = "AED axis: trailblazing (2 replicates) against the final round (20 replicates).")
```

Two entries in the first table are the reason this round was run.

**CoSPAR's top half was noise, not structure.** The trailblazing pass read
0.27, 0.62 and 0.31 at Gini 0.7, 0.8 and 0.9 — a zig-zag that no story
explains and that made "where does CoSPAR fall away" unanswerable. At 20
replicates the same levels read
`r .metric_at("gini", "CoSPAR", 0.7, "spearman_mean")`,
`r .metric_at("gini", "CoSPAR", 0.8, "spearman_mean")` and
`r .metric_at("gini", "CoSPAR", 0.9, "spearman_mean")`: monotone, with the
break between Gini 0.6 and 0.7. Every curve in this round is monotone in its
level, on every method and both axes, which is what a figure of this kind
needs in order to be read as a trend rather than a set of bars.

**CYFER at Gini 0.9 is less bad than two replicates suggested**: 0.52 became
`r .metric_at("gini", "CYFER", 0.9, "spearman_mean")`, with an SD of
`r .metric_at("gini", "CYFER", 0.9, "spearman_sd")` that explains how the
two-replicate estimate landed as low as it did. The interpretation of that
level is in Section 3.4.

The AED axis moved very little between rounds — no entry by more than 0.08 —
which is the expected behaviour when a curve is already smooth, and is a
reassurance about the trailblazing round rather than a new result.

## The fate scores behind the gene calls

```{r score_bars, fig.width=9, fig.height=4.5}
plot(.plot_barplot(pass_list$gini$summary_df[
  pass_list$gini$summary_df$method != "Lineage-DE", ], "gini", "score_cor",
  "cor(fate score, true fate potential) over t1 cells",
  "Gini axis: per-cell fate scores"))
plot(.plot_barplot(pass_list$aed$summary_df[
  pass_list$aed$summary_df$method != "Lineage-DE", ], "aed", "score_cor",
  "cor(fate score, true fate potential) over t1 cells",
  "AED axis: per-cell fate scores"))
```

**Lineage-DE has no bars here because it computes no per-cell fate score**: it
splits clones into a high and a low group by t2 size and compares the two
groups of t1 cells gene by gene, so the only per-cell quantity it touches is
the binary group label. Its absence is a property of the method, not a missing
computation.

CYFER's `cell_imputed_score` correlates 0.85–0.97 with `Z_true` at every level
of both axes. CoSPAR's fate bias — over the t1 cells it actually scores, with
extinct-clone cells excluded — tops out near 0.75 at the easy ends and decays
to `r .metric_at("gini", "CoSPAR", 0.9, "score_cor_mean")` at Gini 0.9 and
`r .metric_at("aed", "CoSPAR", 1, "score_cor_mean")` at AED 1.0.

**The finding worth the paper is in the AED panel.** CYFER's fate score is
essentially flat along that axis —
`r .metric_at("aed", "CYFER", 0.1, "score_cor_mean")` at AED 0.1 and
`r .metric_at("aed", "CYFER", 1, "score_cor_mean")` at AED 1.0, never dipping
below 0.93 — while its gene-level score over the same datasets falls from
`r .metric_at("aed", "CYFER", 0.1, "spearman_mean")` to
`r .metric_at("aed", "CYFER", 1, "spearman_mean")`. Within-clone heterogeneity
degrades **what genes can be read off the fate potential, not the fate
potential itself**. CoSPAR shows no such decoupling: its fate score falls in
step with its genes (0.75 to 0.15), so this is specific to CYFER rather than a
property of the generator at high AED.

The mechanism is the one the metric is built on. The operational truth is the
per-gene Spearman of expression with `Z_true` over the t1 cells; as the
within-clone spread grows, a clone's cells scatter in expression while their
fate potentials stay tied to the clone, so the same fate potential is paired
with increasingly noisy expression and the gene-level signal weakens even
though the per-cell fate potential is recovered as well as ever. For the paper
this separates two claims that are easy to conflate: CYFER's per-cell scores
are robust to heterogeneity, and *gene discovery* from those scores is not.

## Where the top Gini level sits

```{r gini_top_diagnostics}
gini_df <- unique(pass_list$gini$summary_df[, c("level", "t2_num_zero_mean",
                                                "t2_max_clone_mean",
                                                "t2_total_mean",
                                                "num_t1_excluded_cospar_mean")])
gini_df$pct_largest_clone <- round(100 * gini_df$t2_max_clone_mean /
                                     gini_df$t2_total_mean, 1)
rownames(gini_df) <- NULL
knitr::kable(gini_df[, c("level", "t2_num_zero_mean", "t2_max_clone_mean",
                         "pct_largest_clone", "num_t1_excluded_cospar_mean")],
             col.names = c("Gini target", "extinct clones", "largest t2 clone",
                           "% of t2 in it", "t1 cells excluded from CoSPAR"),
             digits = 1,
             caption = "Gini axis: what the top of the ladder does to the data. Means over 20 replicates.")
```

At Gini 0.9 an average of 34 of the 100 clones leave no descendants and the
largest clone holds about half of the 3,000 t2 cells. The operational truth
over 1,000 t1 cells is therefore dominated by the few dozen cells belonging to
one or two clones, and **the level is better described as one where the truth
itself is barely estimable than as one where the methods fail**. CYFER's
per-cell fate score is the evidence: it is still
`r .metric_at("gini", "CYFER", 0.9, "score_cor_mean")` there, so the fate
potentials are recovered; it is the gene-level readout that degrades, for the
same reason as in Section 3.3. The figure caption should say this rather than
leaving the top bar to be read as a failure of the method.

## Secondary metrics

```{r secondary_tables}
knitr::kable(.summary_table(pass_list$gini$summary_df, "Gini target",
                            "pearson"),
             caption = "Gini axis: Pearson version of the metric, mean (SD).")
knitr::kable(.summary_table(pass_list$aed$summary_df, "AED target", "pearson"),
             caption = "AED axis: Pearson version of the metric, mean (SD).")
knitr::kable(.summary_table(pass_list$gini$summary_df, "Gini target",
                            "jaccard"),
             caption = "Gini axis: Jaccard of the called gene sets at BH q < 0.05, mean (SD).")
knitr::kable(.summary_table(pass_list$aed$summary_df, "AED target", "jaccard"),
             caption = "AED axis: Jaccard of the called gene sets at BH q < 0.05, mean (SD).")
```

The Pearson version tracks the Spearman one and is uniformly higher, as it was
in the trailblazing round; nothing in the ordering of the methods depends on
which is used. The Jaccard columns are noisier and should stay secondary:
they depend on a BH threshold applied to p-values from three different
approximations, and at the hard ends of both axes they compare sets that are
nearly empty. One entry is worth not over-reading — lineage-DE's Jaccard at
AED 0.85 (`r .metric_at("aed", "Lineage-DE", 0.85, "jaccard_mean")`) exceeds
its value at easier levels while its Spearman there is
`r .metric_at("aed", "Lineage-DE", 0.85, "spearman_mean")`; a set-overlap
measure can rise as the sets shrink toward the few strongest genes, which is
why the Spearman metric is the headline one.

## Realized statistics and calibration

```{r realized_tables}
.realized_table <- function(summary_df, axis){
  df <- unique(summary_df[, c("level", "gini_t2_mean", "aed_mean_mean",
                              "aed_q05_mean", "aed_q95_mean",
                              "t2_total_mean", "t2_num_zero_mean",
                              "spearman_aed_t2size_mean",
                              "pca_heritability_mean")])
  rownames(df) <- NULL
  colnames(df) <- c("target", "realized Gini", "realized AED", "AED 5th pct",
                    "AED 95th pct", "t2 cells", "extinct clones",
                    "cor(AED, t2 size)", "PCA heritability")
  df
}
knitr::kable(.realized_table(pass_list$gini$summary_df, "gini"), digits = 3,
             caption = "Gini axis: realized statistics, means over 20 replicates.")
knitr::kable(.realized_table(pass_list$aed$summary_df, "aed"), digits = 3,
             caption = "AED axis: realized statistics, means over 20 replicates.")
```

The calibration did its job on both axes. On the AED axis the **held** t2 Gini
realizes 0.48–0.54 at every level, so the axis moves heterogeneity without
dragging clone-size inequality along with it — the point of placing the
clone-varying spread on the non-causal coordinates. Realized mean AED is
within 0.02 of target at every level except the bottom one, where the known
count-noise floor puts 0.1 at 0.124. On the Gini axis the held mean AED stays
at 0.40–0.43 throughout, and the realized Gini tracks the target to within
about 0.05, undershooting slightly in the middle of the ladder (target 0.6
realizes 0.57) — replicate scatter around the calibration, not a calibration
error, and the barplot x-axis carries the realized value beneath each target
so the figure is honest about it.

The spread-size coupling stays within sampling noise of zero at every level of
both axes, as designed: the per-clone AED *variation* is fate-irrelevant by
construction, which is one sentence the Methods owes the reader.

## Convergence and the reproducibility check

```{r convergence, include=FALSE}
conv_df <- rbind(pass_list$gini$summary_df[, c("level", "method",
                                               "num_converged",
                                               "num_replicates")],
                 pass_list$aed$summary_df[, c("level", "method",
                                              "num_converged",
                                              "num_replicates")])
num_full_conv <- sum(conv_df$num_converged == conv_df$num_replicates)
num_conv_rows <- nrow(conv_df)
```

All `r num_full_conv` of the `r num_conv_rows` level-by-method cells have all
20 replicates converged, i.e. every one of the `r num_datasets * 2` datasets:
CYFER's
cross-validated fit on all of them (including the levels where 34 clones are
extinct, which it accepts by design), and CoSPAR's Python pipeline returned a
usable transition map and fate bias on all of them. There is no filtering
anywhere behind these curves.

```{r drift_check, include=FALSE}
.drift_check <- function(axis){
  fin_df <- pass_list[[axis]]$detail_df
  r2_df <- pass_list[[paste0(axis, "_r2")]]$detail_df
  fin_df <- fin_df[fin_df$replicate <= 2, ]
  key_vec <- function(df) paste0(df$level_idx, "_", df$replicate, "_",
                                 df$method)
  r2_df <- r2_df[match(key_vec(fin_df), key_vec(r2_df)), ]
  max(abs(fin_df$spearman - r2_df$spearman), na.rm = TRUE)
}
drift_gini <- .drift_check("gini")
drift_aed <- .drift_check("aed")
num_drift_rows <- sum(pass_list$gini$detail_df$replicate <= 2)
```

The seed scheme is unchanged from the trailblazing round, so replicates 1 and
2 of every level regenerate the `_r2` datasets exactly. They do: across the
`r num_drift_rows` comparable rows per axis the largest absolute difference in
the metric is `r signif(drift_gini, 3)` on the Gini axis and
`r signif(drift_aed, 3)` on the AED axis — identical at the four significant
digits the CSVs carry, CoSPAR's Python shell-out included. The first two of the 20
replicates are therefore a drift check on the whole pipeline, and replicates
3–20 are the new information.

# What this means for Figure 4

Three claims the panels now support, in the form the caption can make them:

1. **CYFER recovers the genes behind expansion across the whole range of
   clone-size inequality and within-clone heterogeneity that real data
   present**, at 0.87–0.95 along the Gini axis through 0.8 and 0.70–0.96 along
   the AED axis, while both comparators degrade well before the hard end. The
   real-data Gini anchors (0.64 at t1, 0.72–0.81 at day 10) sit in the region
   where CoSPAR has already broken and CYFER has not.
2. **The comparators fail in different ways, and neither failure is about the
   coherence prior.** CoSPAR's ceiling at the easy end (about 0.75) is the
   uniform barcode link and the extinct clones, not the assumption that
   transcriptomic neighbours share fate — the trailblazing round settled that
   by showing insensitivity to the clone-specific drift. Lineage-DE's decline
   is the coarsening of a continuous fate potential into two clone groups.
3. **The limits of the hard ends are properties of the data, not of CYFER.**
   At Gini 0.9 a third of clones are extinct and one clone holds half of t2;
   at AED 1.0 clone identity at t1 is spread alone. In both regimes CYFER's
   per-cell fate score stays above 0.85 while gene-level recovery falls, which
   is the honest statement to make.

# What this round does not cover

- The `tau_delta` supplementary row (memo Section 2.4) at the middle levels of
  both axes, deliberately kept out of the final-round driver so it could not
  delay the headline sweeps. The trailblazing round already showed CoSPAR
  moves by less than 0.1 over a four-fold range.
- A formal permutation p-value. The label-permutation check of the
  trailblazing round confirmed the metrics scatter around 0 under permuted
  labels, but CYFER's null spread is wide enough that a p-value would need
  hundreds of permutations.
- The harder 300-weak-gene variant of the gene model (memo Section 9), which
  remains the honest answer to "is a structural set of 100 loading genes too
  easy".
- The top AED level is qualitatively different from the rest of its ladder
  rather than merely harder: at `tau = 0` clones have no centre, so
  lineage-DE's two groups differ in variance and not in mean. The text should
  say so where that level is discussed.

# File inventory

| What | Where |
|---|---|
| Driver | `kevin/Writeup21_new-simulations/sim_final_claude.R` |
| RDS | `OUT_ROOT/Writeup21_new-simulations/sim_{gini,aed}_claude_final.rds` |
| Progress logs | `OUT_ROOT/Writeup21_new-simulations/progress_{gini,aed}_claude_final.txt` |
| CSVs | `csv/kevin/Writeup21/sim_{gini,aed}_{summary,replicate_details,calibration}_final.csv` |
| Figures | `fig/kevin/Writeup21/sim_{gini,aed}_*_final.png` |
| Design | `kevin/Writeup21_new-simulations/simulation_design_claude.md` |
| Trailblazing round | `kevin/Writeup21_new-simulations/Writeup21_trailblazing-report_claude.html` |

`make_csvs_claude.R` and `plot_barplots_claude.R` now take
`WRITEUP21_ROUND=final` to read this round instead of the trailblazing one;
both rounds' files sit side by side.
