---
title: "Writeup21: trailblazing round of the Gini and heterogeneity sweeps"
author: "Kevin Z. Lin (drafted by Claude)"
date: "2026-09-23"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
```

```{r paths}
repo_dir <- file.path("/Users/kevinlin/Library/CloudStorage/Dropbox",
                      "Collaboration-and-People/archive/Nancy/multiomeFate",
                      "git/multiomeFate_analysis")
csv_dir <- file.path(repo_dir, "csv", "kevin", "Writeup21")
```

```{r read_csvs}
.read_pass <- function(axis, suffix){
  list(summary_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_summary", suffix, ".csv")), stringsAsFactors = FALSE),
       detail_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_replicate_details", suffix, ".csv")),
    stringsAsFactors = FALSE),
       calibration_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_calibration", suffix, ".csv")),
    stringsAsFactors = FALSE))
}
pass_list <- list(gini_10 = .read_pass("gini", ""),
                  gini_20 = .read_pass("gini", "_pca20"),
                  gini_r2 = .read_pass("gini", "_r2"),
                  aed_10 = .read_pass("aed", ""),
                  aed_20 = .read_pass("aed", "_pca20"),
                  aed_r2 = .read_pass("aed", "_r2"))
tau_delta_df <- utils::read.csv(file.path(csv_dir,
                                          "check_tau_delta_easy_end_pca20.csv"),
                                stringsAsFactors = FALSE)
permutation_df <- utils::read.csv(file.path(csv_dir,
                                            "check_permutation_r2.csv"),
                                  stringsAsFactors = FALSE)
umap_df <- utils::read.csv(file.path(csv_dir, "umap_coords_r2.csv"),
                           stringsAsFactors = FALSE)
demo_df <- utils::read.csv(file.path(csv_dir, "demo_aed_range_claude.csv"),
                           stringsAsFactors = FALSE)

method_level_vec <- c("CYFER", "CoSPAR", "Lineage-DE")
method_color_vec <- c(CYFER = "#D55E00", CoSPAR = "#0072B2",
                      "Lineage-DE" = "#009E73")
```

# Overview

This report summarizes the **trailblazing round** of the two simulation sweeps
designed in `simulation_design_claude.md` (the design memo) for Figure 4 of the
Nature Methods submission of *Resolution of Selection Versus Adaptation in
Cellular Evolution*. The round ran on the laptop in three passes:

1. **10-PC pass** (2026-09-22): the memo as first written; kept as the record
   of why the embedding dimension matters.
2. **20-PC pass** (2026-09-22, suffix `_pca20`): the variant the first pass
   showed to be necessary (Section 7.7); kept as the record behind the review
   decisions.
3. **`_r2` pass** (2026-09-23): the design as settled after Kevin's review of
   the first two passes — 20 PCs, the Gini ladder re-spaced to
   {0.3, ..., 0.9}, t1 cells of extinct clones excluded from CoSPAR, the
   split-half reliability dropped — plus the Section 7 label-permutation check
   and UMAPs of example datasets. **All headline results below are this
   pass.**

Every number below is read from the CSVs in `csv/kevin/Writeup21/`; the
figures in `fig/kevin/Writeup21/` are re-drawn here from the same tables.

| Stage | What it does | Where |
|---|---|---|
| Generate | Hierarchical latent states, one causal coordinate, negative-binomial counts for t1 cells and their t2 children | `func_generate_claude.R` |
| Calibrate | `h²` from the AED target, `s` bisected for the Gini target, `h²` re-centred on the realized AED | `calibrate_scale()`, `run_sweep()` |
| Embed | PCA on log-normalized t1 counts, t2 projected | `compute_embedding()` |
| Fit | CYFER, lineage-DE, CoSPAR, each to a per-gene association vector | `func_methods_claude.R` |
| Score | Spearman of each vector with the operational truth over 2,000 genes | `score_gene_stats()` |
| Sweep | 7 levels × 2 replicates per axis, progress file, RDS per level | `sim_gini_claude.R`, `sim_aed_claude.R` |
| Report | CSVs and barplots | `make_csvs_claude.R`, `plot_barplots_claude.R` |

# Goal

Figure 4 of the paper is planned as two panels, each a ladder of seven simulated
settings, comparing three ways of finding the genes associated with clonal
expansion from a barcoded two-time-point experiment:

- **4A, varying clone-size skewness.** As the Gini index of the later-time-point
  (t2) clone sizes rises from "many clones expanded" to "one or two clones hold
  everything", how well does each method recover the expansion genes?
- **4B, varying intraclonal heterogeneity.** As the within-clone average squared
  Euclidean distance (AED) in the embedding rises from "every cell in a clone
  looks alike" to "clone membership says nothing about a cell's state", the same
  question.

The story both panels must tell is a **gradient**: at the easy end of each axis
all three methods do well; at the hard end only CYFER should, because it models
the expansion potential of every cell rather than of clones (lineage-DE) or of
smooth state transitions tied to clone-level barcode links (CoSPAR). The
trailblazing round has a narrower goal than the figure itself: to run both
sweeps end to end at the held fixed values (t2 Gini 0.5 during the AED sweep,
mean squared AED 0.4 during the Gini sweep, clone-specific t2 drift
`τ_δ = 0.6`) with two replicates per level, and to read off (i) whether the
pipeline works, (ii) where along each axis the comparators fall away, so the
final round can confirm its fixed values and level spacing, and (iii) whether
the two initially provisional design settings
(`spread_variation = "noncausal"` and `τ_δ = 0.6`) pan out. All three
questions are answered below.

# How the data are generated

One generator serves both sweeps (memo Section 2). Sizes: 100 clones, 10 t1
cells per clone, 3,000 expected t2 cells with no cap on clone size and extinct
clones kept, 2,000 genes, latent dimension `d = 10`.

**Latent state at t1.** Each clone `l` has a centre `m_l ~ N(0, τ² I)` and each
of its cells a deviation with per-coordinate standard deviation `σ_lk`. The
within-clone spread varies across clones through a multiplier
`g_l ~ Gamma(κ, κ)` with `κ = 1.5`, applied to the nine **non-causal**
coordinates only (`spread_variation = "noncausal"`), so a clone's spread on the
expansion axis, and hence its expected t2 size, is the same for tight and
diffuse clones. The two knobs are a scale and a share: `τ = s · sqrt(h²)`,
`σ_w = s · sqrt(1 − h²)`.

**Fate potential.** The first coordinate is the expansion axis:
`Z_i = β_0 + β s_i1` with `β = 1.5`, `β_0` solved so the expected number of t2
cells is 3,000, and each t1 cell leaves `Poisson(exp(Z_i))` children. The t2
size of a clone is the sum over its cells, zeros included.

**Genes.** 100 expansion genes load on the causal coordinate (magnitudes
`Unif(0.5, 1)`, half positive, half negative); every gene loads on the other
nine coordinates with dense `N(0, 0.25²)` loadings. Counts are negative
binomial (`θ = 10`) with log-normal library sizes around 5,000 and `N(0, 1)`
log-baselines.

**The later time point.** The t2 population is the multiset of children. A
child's latent state is `ρ · s_parent + (1 − ρ) · m_l + sqrt(1 − ρ²) σ_l ∘ e +
δ + δ_l` with `ρ = 0.8`, a shared shift `δ` of norm `3s` on the non-causal
coordinates, and a clone-specific shift `δ_l ~ N(0, τ_δ² I)` with `τ_δ = 0.6`.
Only CoSPAR ever sees t2 expression.

**Seeds.** Replicate `r` of level `j` uses seed `seed_base + 100 j + r`
(`seed_base` 1000 for the Gini axis, 2000 for the AED axis); the generator's
three stochastic stages use `10 · seed + {0, 1, 2}`; the calibration draws use
`seed_base + 100 j + 50 + k`. (The first version of the code reused the
calibration draws as replicates and shared sub-streams between consecutive
replicates; see Section 8.)

# UMAPs of example datasets

To give a sense of what the generated data look like, six datasets were
re-generated exactly (from the `_r2` calibration tables and the sweep's seed
scheme; `make_umaps_claude.R`) and embedded by log-normalization, the shared
20-PC PCA, then UMAP on all cells: the easiest, middle and hardest level of
each axis (Gini 0.3 / 0.6 / 0.9 at fixed AED 0.4; AED 0.1 / 0.55 / 1.0 at
fixed Gini 0.5), replicate 1 of each. Three colorings per dataset: time point
(all cells), clone (t1 cells only, one color per clone, legend suppressed),
and true fate potential `Z_true` (t1 cells only).

```{r umap_helpers}
.umap_axis_df <- function(axis){
  sub_df <- umap_df[umap_df$axis == axis, ]
  sub_df$facet <- factor(paste0(sub_df$difficulty, " (", sub_df$level, ")"),
                         levels = paste0(c("easiest", "middle", "hardest"),
                                         " (", sub_df$level[
                                           match(c("easiest", "middle",
                                                   "hardest"),
                                                 sub_df$difficulty)], ")"))
  sub_df
}

.plot_umap_time <- function(axis, title){
  sub_df <- .umap_axis_df(axis)
  # Draw the (many) t2 cells beneath the t1 cells so t1 stays visible.
  sub_df <- sub_df[order(sub_df$time_info, decreasing = TRUE), ]
  plot1 <- ggplot2::ggplot(sub_df, ggplot2::aes(x = umap1, y = umap2,
                                                color = time_info))
  plot1 <- plot1 + ggplot2::geom_point(size = 0.3, alpha = 0.5)
  plot1 <- plot1 + ggplot2::scale_color_manual(
    values = c(t1 = "#D55E00", t2 = "gray60"))
  plot1 <- plot1 + ggplot2::facet_wrap(~ facet, nrow = 1, scales = "free")
  plot1 <- plot1 + ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(size = 2, alpha = 1)))
  plot1 <- plot1 + ggplot2::labs(title = title, color = "time point")
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot1
}

.plot_umap_clone <- function(axis, title){
  sub_df <- .umap_axis_df(axis)
  sub_df <- sub_df[sub_df$time_info == "t1", ]
  plot1 <- ggplot2::ggplot(sub_df, ggplot2::aes(x = umap1, y = umap2,
                                                color = clone_id))
  plot1 <- plot1 + ggplot2::geom_point(size = 0.4, alpha = 0.7)
  plot1 <- plot1 + ggplot2::facet_wrap(~ facet, nrow = 1, scales = "free")
  plot1 <- plot1 + ggplot2::labs(title = title)
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "none")
  plot1
}

.plot_umap_fate <- function(axis, title){
  sub_df <- .umap_axis_df(axis)
  sub_df <- sub_df[sub_df$time_info == "t1", ]
  plot1 <- ggplot2::ggplot(sub_df, ggplot2::aes(x = umap1, y = umap2,
                                                color = z_true))
  plot1 <- plot1 + ggplot2::geom_point(size = 0.4, alpha = 0.8)
  plot1 <- plot1 + ggplot2::scale_color_viridis_c()
  plot1 <- plot1 + ggplot2::facet_wrap(~ facet, nrow = 1, scales = "free")
  plot1 <- plot1 + ggplot2::labs(title = title, color = "Z_true")
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot1
}
```

## Gini axis (fixed AED 0.4)

```{r umap_gini, fig.width=10, fig.height=4}
plot(.plot_umap_time("gini", "Gini axis: all cells by time point"))
plot(.plot_umap_clone("gini", "Gini axis: t1 cells by clone (no legend)"))
plot(.plot_umap_fate("gini", "Gini axis: t1 cells by true fate potential"))
```

## AED axis (fixed Gini 0.5)

```{r umap_aed, fig.width=10, fig.height=4}
plot(.plot_umap_time("aed", "AED axis: all cells by time point"))
plot(.plot_umap_clone("aed", "AED axis: t1 cells by clone (no legend)"))
plot(.plot_umap_fate("aed", "AED axis: t1 cells by true fate potential"))
```

What to look for. The t1 and t2 populations barely overlap (the shared shift
`δ` of about three t1 standard deviations on the non-causal coordinates). On
the Gini axis the clone structure at t1 looks similar at every level — the
axis moves the *fate potential's* spread, not the geometry — while the fate
coloring becomes more extreme toward Gini 0.9. On the AED axis the t1 clones
go from tight patches (AED 0.1) to fully mixed (AED 1.0, where clones have no
centre), while the fate potential stays a smooth gradient because the causal
coordinate keeps the same share-free spread; the fate gradient is carried by
one of 20 PCs, so it is not always aligned with the UMAP's dominant
directions.

# How the simulation suite is set up

## The two axes and their calibration

Both axes are stated in terms of a **realized** statistic, not a generator
parameter:

- **Gini axis:** the t2-only Gini of the 100 clone sizes, zeros included
  (`gini_coef()`, identical to `dineq::gini.wtd`), targets
  `{0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9}`, mean squared AED held at 0.4. The
  first two passes ran `{0.2, 0.3, 0.4, 0.5, 0.6, 0.75, 0.9}`; the `_r2`
  ladder drops 0.2 (unreachable at the fixed AED, Section 7 of the round-1
  report) and is denser between 0.5 and 0.9, where the comparators fall away.
- **AED axis:** the mean over clones of the per-clone squared AED
  (`SqDist_l / SqDist_random` in the shared embedding, computed as twice the
  sum of per-coordinate variances), targets
  `{0.1, 0.25, 0.4, 0.55, 0.7, 0.85, 1.0}`, t2 Gini held at 0.5.

Calibration per level: `h² = 1 − AED target`; `s` bisected (geometric, mean
Gini over 10 latent-only draws, tolerance 0.01); then one full dataset is
generated and embedded, its realized mean AED compared with the target, and
`h²` shifted by the miss up to three times (tolerance 0.03), re-bisecting `s`
each time. The re-centring step was added after the code review (Section 6);
the 10-PC pass ran without it.

```{r calibration_tables}
.calib_table <- function(df, axis_label){
  out_df <- df[, c("level", "h2_initial", "h2", "num_recentre", "aed_check",
                   "latent_scale", "tau", "sigma_w", "gini_calibrated")]
  out_df[sapply(out_df, is.numeric)] <- lapply(out_df[sapply(out_df, is.numeric)],
                                               round, 3)
  colnames(out_df) <- c(axis_label, "h2 initial", "h2 final", "re-centrings",
                        "check AED", "s", "tau", "sigma_w", "Gini (10 draws)")
  out_df
}
knitr::kable(.calib_table(pass_list$gini_r2$calibration_df, "Gini target"),
             caption = "Calibration of the Gini axis (`_r2` pass).")
knitr::kable(.calib_table(pass_list$aed_r2$calibration_df, "AED target"),
             caption = "Calibration of the AED axis (`_r2` pass).")
```

Two things the tables show. The calibrated latent scale `s` runs from 0.37 to
2.06 along the Gini axis: the Gini is moved by making the fate potential
`β s_i1` more or less spread across cells, so the top Gini level is also the
level with the strongest expression signal and the bottom level the weakest.
And the re-centring only hits its wall at the bottom of the AED axis: the
AED-0.1 level needs `h² = 1` (no within-clone latent spread at all) and still
realizes 0.13, the count-noise floor. The re-spaced Gini ladder avoids the
symmetric wall — its dropped 0.2 level would have needed `s = 0.21`, where
even `h² = 1` left the realized AED at 0.55 — and every level of both axes
now checks within 0.03 of its AED target at calibration time.

## Runs

Three passes, each 7 levels × 2 replicates × 2 axes = 28 datasets:

| Pass | Embedding | Gini ladder | CoSPAR extinct-clone t1 cells | Output suffix | Role |
|---|---|---|---|---|---|
| 1 | 10 PCs (memo as first written) | 0.2–0.9 | kept, bias 0.5 | none | the record of why the embedding dimension matters |
| 2 | 20 PCs | 0.2–0.9 | kept, bias 0.5 | `_pca20` | the record behind the review decisions |
| 3 | 20 PCs | 0.3–0.9 re-spaced | **excluded before CoSPAR** | `_r2` | the settled design; every headline number |

All three passes re-centre the AED calibration and use the corrected seed
scheme except pass 1, which predates both. One dataset takes about a minute
(generation 1 s, embedding 5 s, CYFER 3–6 s, lineage-DE 1 s, CoSPAR 15–45 s);
a sweep is 12–14 minutes per axis and the two axes run concurrently.

# Methods compared

All three methods see the same 1,000 t1 cells, the same clone labels, the same
t2 clone sizes and, where they use one, the same embedding (PCA of
log-normalized t1 counts; t2 cells projected with the t1 centre and rotation).

**CYFER.** `cyfer()` with 3-fold cross-validation over a 10-value λ path and
`cyfer_finalize()` on the PCs, with the zero-count clones **kept** (the
sim1–sim7 scripts drop them; here they carry the signal at high Gini). The
per-gene statistic is the Spearman correlation of each gene's log-normalized
expression with `cell_imputed_score` over t1 cells.

**Lineage-DE.** Clones are split into "high" (t2 size above the *mean*) and
"low"; every gene is compared between the t1 cells of high and low clones by
the Wilcoxon rank-sum test, and the statistic is the rank-biserial correlation
`2U / (n_H n_L) − 1`, computed for all genes at once from ranks and
cross-checked against `stats::wilcox.test()` on five genes per dataset (U and
p-value agree to 1e-6).

**CoSPAR.** The matched route: `state_info` is `"t1"` for t1 cells and
`"High"` / `"Low"` for t2 cells by the same mean split;
`infer_Tmap_from_multitime_clones()` on the shared PCA (paper defaults:
`smooth 20 15 10 5`, `sparsity 0.1`, `intraclone 0.2`); `fate_bias(High, Low)`
on the intraclone map; the statistic is the Spearman correlation of each gene
with the fate bias over t1 cells. **In the `_r2` pass the t1 cells of extinct
clones are excluded before CoSPAR runs**: they are single-time clones to
CoSPAR, outside its transition map, and the 0.5 bias it fills in for them
would rank these lowest-fate cells in the middle. Their bias is `NA` instead,
and both the gene correlation and the fate-score diagnostic run over the
included t1 cells only (34–40 % of t1 cells are excluded at Gini 0.9, none at
the easy end). CoSPAR runs in its own conda environment through the
`run-cospar` flat-file templates; `run_cospar.py` is now heavily commented
for manual review of every CoSPAR call.

**The metric.** The truth is operational: `truth_g = Spearman(x_g, Z_true)`
over the same t1 cells. The score of a method is
`Spearman(m, truth)` over all 2,000 genes (Pearson and a Jaccard of BH
`q < 0.05` sets are secondary columns). Per dataset the tables also record the
correlation of each method's per-cell fate score (`Z_hat`, fate bias) with
`Z_true`, so a gene-level failure can be traced to the score or to the gene
step.

# Results

## The headline curves (`_r2` pass)

```{r plot_helpers}
.level_label_vec <- function(summary_df, axis){
  level_df <- unique(summary_df[, c("level_idx", "level", "gini_t2_mean",
                                    "aed_mean_mean", "aed_q05_mean",
                                    "aed_q95_mean")])
  level_df <- level_df[order(level_df$level_idx), ]
  if(axis == "gini"){
    label_vec <- paste0(level_df$level, "\n(",
                        sprintf("%.2f", level_df$gini_t2_mean), ")")
  } else {
    label_vec <- paste0(level_df$level, "\n(",
                        sprintf("%.2f", level_df$aed_mean_mean), ")\n[",
                        sprintf("%.2f", level_df$aed_q05_mean), ", ",
                        sprintf("%.2f", level_df$aed_q95_mean), "]")
  }
  names(label_vec) <- level_df$level_idx
  label_vec
}

.plot_barplot <- function(summary_df, detail_df, axis, metric, ylab, title){
  summary_df <- summary_df[summary_df$method %in% method_level_vec, ]
  detail_df <- detail_df[detail_df$method %in% method_level_vec, ]
  summary_df$method <- factor(summary_df$method, levels = method_level_vec)
  detail_df$method <- factor(detail_df$method, levels = method_level_vec)
  summary_df$level_factor <- factor(summary_df$level_idx)
  detail_df$level_factor <- factor(detail_df$level_idx)
  mean_col <- paste0(metric, "_mean")
  label_vec <- .level_label_vec(summary_df, axis)
  x_label <- if(axis == "gini"){
    "Gini of t2 clone sizes: target (realized mean)"
  } else {
    "Mean squared AED: target (realized mean) [per-clone 5th-95th pct]"
  }

  plot1 <- ggplot2::ggplot(summary_df,
                           ggplot2::aes(x = level_factor, y = .data[[mean_col]],
                                        fill = method))
  plot1 <- plot1 + ggplot2::geom_col(position = ggplot2::position_dodge(0.8),
                                     width = 0.7)
  plot1 <- plot1 + ggplot2::geom_point(
    data = detail_df,
    ggplot2::aes(x = level_factor, y = .data[[metric]], group = method),
    position = ggplot2::position_dodge(0.8), size = 1.2, color = "black",
    inherit.aes = FALSE)
  plot1 <- plot1 + ggplot2::geom_hline(yintercept = 0, linewidth = 0.3)
  plot1 <- plot1 + ggplot2::scale_fill_manual(values = method_color_vec)
  plot1 <- plot1 + ggplot2::scale_x_discrete(labels = label_vec)
  plot1 <- plot1 + ggplot2::labs(x = x_label, y = ylab, fill = NULL,
                                 title = title)
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot1
}

.summary_table <- function(summary_df, axis_label){
  wide_df <- data.frame(level = sort(unique(summary_df$level)))
  for(method in method_level_vec){
    sub_df <- summary_df[summary_df$method == method, ]
    sub_df <- sub_df[match(wide_df$level, sub_df$level), ]
    wide_df[[method]] <- sprintf("%.2f (%.2f)", sub_df$spearman_mean,
                                 sub_df$spearman_sd)
  }
  colnames(wide_df)[1] <- axis_label
  rownames(wide_df) <- NULL
  wide_df
}
```

```{r gini_r2_bar, fig.width=9, fig.height=5}
plot(.plot_barplot(pass_list$gini_r2$summary_df, pass_list$gini_r2$detail_df,
                   "gini", "spearman", "Spearman(method, truth) over genes",
                   "4A: Gini axis, `_r2` pass"))
```

```{r aed_r2_bar, fig.width=9, fig.height=5}
plot(.plot_barplot(pass_list$aed_r2$summary_df, pass_list$aed_r2$detail_df,
                   "aed", "spearman", "Spearman(method, truth) over genes",
                   "4B: AED axis, `_r2` pass"))
```

```{r summary_tables_r2}
knitr::kable(.summary_table(pass_list$gini_r2$summary_df, "Gini target"),
             caption = "Gini axis, `_r2` pass: Spearman of each method's per-gene vector with the truth, mean (SD) over 2 replicates.")
knitr::kable(.summary_table(pass_list$aed_r2$summary_df, "AED target"),
             caption = "AED axis, `_r2` pass: Spearman of each method's per-gene vector with the truth, mean (SD) over 2 replicates.")
```

**Both panels show the intended gradient.** On the Gini axis CYFER stays at
0.90–0.93 from Gini 0.3 through 0.8 while lineage-DE holds at 0.75–0.78
through 0.6 and then degrades (0.48–0.67, with replicate SDs up to 0.35), and
CoSPAR — now with the extinct-clone exclusion — holds near 0.70 through 0.6
and then becomes erratic (0.27, 0.62 and 0.31 at 0.7, 0.8 and 0.9, SDs up to
0.31). On the AED axis CYFER stays at 0.79–0.97 across the whole ladder while
CoSPAR falls from 0.69–0.76 at the easy end to −0.03 at AED 1.0, and
lineage-DE from 0.75–0.84 to 0.23. CYFER is the best method at every level of
both axes except Gini 0.9, where all three land in an uninformative 0.3–0.6
band.

Three qualifications. At the top Gini level (0.9) CYFER's gene-level score
drops to 0.52 (SD 0.05) even though its fate score still correlates 0.82 with
the truth: one clone holds about 1,100 of the 3,000 t2 cells and 25 clones
are extinct, so the operational truth over 1,000 cells is itself dominated by
a few dozen cells. Second, the extinct-clone exclusion moves CoSPAR's
fall-off point: with the 0.5-filled cells of the earlier passes CoSPAR
dropped below 0.3 already at Gini 0.6, whereas with the exclusion it holds
near 0.7 there, so **the "last easy level" of the Gini axis now reads 0.6**
(all three ≥ 0.70) rather than the held 0.5; with two replicates that
difference is within noise, and whether to move the fixed Gini from 0.5 to
0.6 for the final round is an open decision (memo Sections 8 and 9). The AED
axis is unchanged: the last easy level is 0.4 (CoSPAR 0.47 and lineage-DE
0.51 at 0.55), the held value — and there the exclusion moves CoSPAR by at
most +0.02 (CYFER and lineage-DE reproduce the earlier pass to four
decimals, same seeds), so CoSPAR's fall-off along the AED axis is genuine
rather than extinct-clone bookkeeping. Third, everything above Gini 0.6 rests on two
replicates with large comparator SDs and should not be over-read before the
20-replicate round.

## The fate scores behind the gene calls

```{r score_bars, fig.width=9, fig.height=4.5}
plot(.plot_barplot(pass_list$gini_r2$summary_df, pass_list$gini_r2$detail_df,
                   "gini", "score_cor",
                   "cor(fate score, true fate potential) over t1 cells",
                   "Gini axis, `_r2` pass: per-cell fate scores"))
plot(.plot_barplot(pass_list$aed_r2$summary_df, pass_list$aed_r2$detail_df,
                   "aed", "score_cor",
                   "cor(fate score, true fate potential) over t1 cells",
                   "AED axis, `_r2` pass: per-cell fate scores"))
```

CYFER's `cell_imputed_score` correlates 0.82–0.98 with `Z_true` at every level
of both axes (0.82 at Gini 0.9, at least 0.93 everywhere else). CoSPAR's fate
bias — computed over the t1 cells it now actually scores, extinct-clone cells
excluded — tops out near 0.7 at the easy end of both axes and decays to 0.14
at AED 1.0 and 0.40 at Gini 0.9; its ceiling is
not the t2 drift (Section 7.6). **Lineage-DE has no bars in these plots
because it does not compute a per-cell fate score at all**: it splits the
clones into a high and a low group by their t2 sizes and compares the two
groups of t1 cells gene by gene, so the only per-cell quantity it ever touches
is the binary group label, and there is no natural per-cell number to
correlate with `Z_true`. (One could assign every cell its clone's t2 size,
but that is a clone-level score, not a per-cell one, and it is not part of
the lineage-DE recipe.) Its absence here is a property of the method, not a
missing computation.

## Secondary metrics

```{r secondary_tables}
.secondary_table <- function(summary_df, axis_label){
  out_df <- summary_df[order(summary_df$level, summary_df$method),
                       c("level", "method", "pearson_mean", "jaccard_mean",
                         "num_converged")]
  out_df$pearson_mean <- round(out_df$pearson_mean, 2)
  out_df$jaccard_mean <- round(out_df$jaccard_mean, 2)
  colnames(out_df) <- c(axis_label, "method", "Pearson", "Jaccard (q < 0.05)",
                        "converged / 2")
  rownames(out_df) <- NULL
  out_df
}
knitr::kable(.secondary_table(pass_list$gini_r2$summary_df, "Gini target"),
             caption = "Gini axis, `_r2` pass: secondary metrics, mean over 2 replicates.")
knitr::kable(.secondary_table(pass_list$aed_r2$summary_df, "AED target"),
             caption = "AED axis, `_r2` pass: secondary metrics, mean over 2 replicates.")
```

Every CYFER fit and every CoSPAR run converged (84 datasets across the three
passes, plus the permutation and `τ_δ` checks; no `NULL` fit and no collapsed
bias). The Jaccard column orders the
methods the same way as the Spearman headline at almost every level but is
noisier, as expected from a thresholded set.

## Realized statistics and diagnostics

One row per level, averaged over the 2 replicates. (The realized statistics
are dataset-level, so the three method rows of the summary table collapse to
one row here; an earlier version of this table let the data frame's original
row names — 1, 4, 7, ... , the positions of the first method row of each
level — leak into the display, which meant nothing and is now suppressed.)

```{r realized_table}
.realized_table <- function(summary_df, axis_label){
  out_df <- unique(summary_df[, c("level", "gini_t2_mean", "aed_mean_mean",
                                  "aed_q05_mean", "aed_q95_mean",
                                  "t2_max_clone_mean", "t2_num_zero_mean",
                                  "num_high_clones_mean",
                                  "num_t1_excluded_cospar_mean",
                                  "spearman_aed_t2size_mean",
                                  "pca_heritability_mean")])
  out_df <- out_df[order(out_df$level), ]
  out_df[sapply(out_df, is.numeric)] <- lapply(out_df[sapply(out_df, is.numeric)],
                                               round, 2)
  colnames(out_df) <- c(axis_label, "Gini t2", "mean AED", "AED 5th pct",
                        "AED 95th pct", "largest clone", "extinct clones",
                        "high clones", "t1 cells excluded (CoSPAR)",
                        "Spearman(AED_l, Y_l)", "PCA heritability")
  rownames(out_df) <- NULL
  out_df
}
knitr::kable(.realized_table(pass_list$gini_r2$summary_df, "Gini target"),
             caption = "Gini axis, `_r2` pass: realized statistics per level, mean over 2 replicates.")
knitr::kable(.realized_table(pass_list$aed_r2$summary_df, "AED target"),
             caption = "AED axis, `_r2` pass: realized statistics per level, mean over 2 replicates.")
```

- **The calibration holds** up to replicate scatter: the realized mean AED
  during the Gini sweep is 0.40–0.43 at every level, the realized Gini during
  the AED sweep is 0.45–0.55, and the realized Gini along its own axis is
  within 0.06 of target in the mean of two replicates (levels 0.4 and 0.8
  realize 0.36 and 0.74), consistent with the ±0.05 per-replicate scatter of
  memo Section 4.2. The two floors are as expected: AED 0.1 realizes 0.12 and
  AED 1.0 realizes 0.94.
- **The per-clone AED spans the intended range**: at the top AED level the
  5th–95th percentiles are 0.27–1.71 (the per-clone maximum exceeds 2).
- **The spread-size coupling is gone**, as the `"noncausal"` placement was meant
  to achieve: Spearman between a clone's AED and its t2 size is between −0.10
  and 0.12 at every level of both axes.
- **The Gini axis reaches the intended regime**: at Gini 0.9 the largest clone
  holds about 1,100 of 3,000 t2 cells and 25 clones are extinct; the
  lineage-DE high group shrinks from 42 clones at Gini 0.3 to about 12, and
  250 of the 1,000 t1 cells are excluded from CoSPAR as extinct-clone cells
  (none are at the easy end of either axis).

## Sanity checks (memo Section 7)

The memo's Section 7 checklist, item by item. All but the label-permutation
check are columns of the replicate tables, summarized above.

- **Label permutation** (`check_permutation_claude.R`): at the middle level of
  each axis (Gini 0.6, AED 0.55), the t1 cells' clone labels were permuted
  before fitting — the t2 sizes stay attached to the clone names, so every
  method sees the same clone-size distribution with the feature–fate link
  severed — and all three methods were re-run on five fresh permutations per
  axis:

```{r permutation_table}
perm_key_df <- unique(permutation_df[, c("axis", "method")])
perm_summary_list <- lapply(seq_len(nrow(perm_key_df)), function(i){
  sub_df <- permutation_df[permutation_df$axis == perm_key_df$axis[i] &
                             permutation_df$method == perm_key_df$method[i], ]
  data.frame(axis = perm_key_df$axis[i],
             method = perm_key_df$method[i],
             num_perm = nrow(sub_df),
             spearman_mean = round(mean(sub_df$spearman), 3),
             spearman_min = round(min(sub_df$spearman), 3),
             spearman_max = round(max(sub_df$spearman), 3),
             score_cor_min = round(suppressWarnings(min(sub_df$score_cor,
                                                        na.rm = TRUE)), 3),
             score_cor_max = round(suppressWarnings(max(sub_df$score_cor,
                                                        na.rm = TRUE)), 3),
             stringsAsFactors = FALSE)
})
perm_summary_df <- do.call(rbind, perm_summary_list)
perm_summary_df[perm_summary_df$method == "Lineage-DE",
                c("score_cor_min", "score_cor_max")] <- NA
colnames(perm_summary_df) <- c("axis", "method", "permutations",
                               "Spearman mean", "Spearman min", "Spearman max",
                               "cor(score, Z_true) min",
                               "cor(score, Z_true) max")
rownames(perm_summary_df) <- NULL
knitr::kable(perm_summary_df,
             caption = "Label-permutation check, summarized over the permutations of each axis.")
```

  The permuted metrics scatter around 0 with no consistent sign — compare
  0.88–0.97 (CYFER), 0.68–0.76 (CoSPAR) and 0.75–0.84 (lineage-DE) at the
  same levels with intact labels. One honest caveat: CYFER's null spread is
  the widest of the three. Under permuted labels its cross-validation does
  not shrink the fit all the way to a constant, so the permuted score is an
  overfit *random* direction in the 20-PC space; a random direction's chance
  alignment with `Z_true` has standard deviation near `1/sqrt(20) ≈ 0.22`,
  and the gene-level Spearman amplifies whatever alignment a draw happens to
  have (the most extreme permutation reached a gene-level 0.61 from a chance
  score alignment of 0.39). The check is qualitative — it confirms no
  systematic leak, but a formal permutation *p*-value would need hundreds of
  permutations, which is final-round material if wanted.

- **Realized statistics within tolerance**: Section 7.4; the calibration holds
  everywhere except the count-noise floor at the bottom of the AED axis.
- **Spread-size coupling** near 0 under `"noncausal"`: the
  `Spearman(AED_l, Y_l)` column of Section 7.4.
- **CYFER convergence** and **CoSPAR group sizes / collapsed-bias detection**:
  the `converged / 2` column of Section 7.3 (a constant CoSPAR bias would be
  reported as a failed run; none occurred).
- **t1 cells excluded from CoSPAR** per level: Section 7.4.
- **PCA heritability** per level, for cross-reference to sim3: Section 7.4.
- **Split-half reliability** was on the memo's original list and has been
  dropped by decision (its Spearman form is 0.2–0.7 by construction with
  1,900 null genes, so it cannot serve as the intended "well below 0.9" flag).

## Does the clone-specific t2 drift explain CoSPAR's ceiling?

If CoSPAR did poorly even at the easy end because of the clone-specific
drift, `τ_δ` would be the thing to lower. At the easiest AED level
(`h² = 0.9`, Gini 0.5, two seeds, 20 PCs; run during the `_pca20` pass, so
before the extinct-clone exclusion, which touches few cells at this easy
level) CoSPAR was re-run at `τ_δ ∈ {0, 0.3, 0.6, 1.2}`:

```{r tau_delta_table}
td_df <- tau_delta_df[tau_delta_df$method == "CoSPAR", ]
td_wide_df <- data.frame(tau_delta = sort(unique(td_df$tau_delta)))
for(seed in sort(unique(td_df$seed))){
  sub_df <- td_df[td_df$seed == seed, ]
  sub_df <- sub_df[match(td_wide_df$tau_delta, sub_df$tau_delta), ]
  td_wide_df[[paste0("Spearman, seed ", seed)]] <- round(sub_df$spearman, 2)
  td_wide_df[[paste0("cor(bias, Z_true), seed ", seed)]] <- round(sub_df$score_cor, 2)
}
knitr::kable(td_wide_df,
             caption = "CoSPAR at the easiest AED level as the clone-specific t2 drift varies.")
```

CoSPAR is essentially insensitive to `τ_δ`: its fate-score correlation moves
from 0.72 to 0.65–0.67 over a four-fold range of drift, and its gene-level score
by about the same. Its ceiling of roughly 0.7 on these data is therefore the
uniform barcode link (every t1 cell of a clone tied equally to every t2 cell of
that clone), not the loss of cross-clone coherence at t2. `τ_δ = 0.6` can
stand, and the paper's framing should say "uniform link and extinct clones"
rather than "smooth continuum" (now recorded as settled in memo Section 9).

## The 10-PC pass: why the embedding dimension matters

This subsection compares the two 2026-09-22 passes (10 PCs versus 20 PCs, both
on the original Gini ladder and before the CoSPAR exclusion) and is kept as
the record behind the decision to adopt 20 PCs.

```{r compare_passes, fig.width=9, fig.height=4.5}
.compare_df <- function(axis){
  a_df <- pass_list[[paste0(axis, "_10")]]$summary_df
  b_df <- pass_list[[paste0(axis, "_20")]]$summary_df
  a_df$pass <- "10 PCs"
  b_df$pass <- "20 PCs"
  out_df <- rbind(a_df[, c("level", "method", "pass", "spearman_mean",
                           "score_cor_mean")],
                  b_df[, c("level", "method", "pass", "spearman_mean",
                           "score_cor_mean")])
  out_df$method <- factor(out_df$method, levels = method_level_vec)
  out_df
}
for(axis in c("gini", "aed")){
  cmp_df <- .compare_df(axis)
  plot1 <- ggplot2::ggplot(cmp_df, ggplot2::aes(x = level, y = spearman_mean,
                                                color = method,
                                                linetype = pass))
  plot1 <- plot1 + ggplot2::geom_line()
  plot1 <- plot1 + ggplot2::geom_point()
  plot1 <- plot1 + ggplot2::scale_color_manual(values = method_color_vec)
  plot1 <- plot1 + ggplot2::labs(
    x = if(axis == "gini") "Gini target" else "Mean squared AED target",
    y = "Spearman(method, truth), mean of 2", color = NULL, linetype = NULL,
    title = paste0(if(axis == "gini") "Gini" else "AED",
                   " axis: 10-PC pass (memo as written) versus 20-PC pass"))
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot(plot1)
}
```

```{r compare_table}
cmp_list <- lapply(c("gini", "aed"), function(axis){
  cmp_df <- .compare_df(axis)
  cmp_df <- cmp_df[cmp_df$method == "CYFER", ]
  wide_df <- data.frame(axis = axis, level = sort(unique(cmp_df$level)))
  for(pass in c("10 PCs", "20 PCs")){
    sub_df <- cmp_df[cmp_df$pass == pass, ]
    sub_df <- sub_df[match(wide_df$level, sub_df$level), ]
    wide_df[[paste0("genes, ", pass)]] <- round(sub_df$spearman_mean, 2)
    wide_df[[paste0("fate score, ", pass)]] <- round(sub_df$score_cor_mean, 2)
  }
  wide_df
})
knitr::kable(do.call(rbind, cmp_list),
             caption = "CYFER under the two embeddings: gene-level Spearman and fate-score correlation. (The 10-PC pass predates the re-centring and seed fixes, so its levels are not identical datasets.)")
```

With the memo's 10-PC embedding CYFER collapses at the hard end of both axes
(gene-level 0.06 at Gini 0.9 and 0.09 at AED 1.0, fate-score correlation 0.06
and 0.01) and at the weak-signal bottom of the Gini axis (0.44), and
lineage-DE, which never uses the embedding, looks best at the top. The
diagnosis, established with side experiments during the run:

- Fitting CYFER on the **true latent coordinates** instead of the PCs gives a
  fate-score correlation of 1.00 at every setting, so the model is not the
  problem.
- The causal axis loads on 100 genes with squared loadings around 0.58, while
  each of the nine non-causal coordinates loads on all 2,000 genes with squared
  loadings 0.0625, so in gene space the causal axis carries about half the
  variance of any non-causal one and is always the **weakest latent
  direction**. At the middle of either axis it sits on PC 10 (correlation
  0.94 with the causal coordinate); at the corners it drops to PC 11
  (correlation 0.83–0.96 with PC 11, at most 0.31 with any of PCs 1–10).
- The R² of `Z_true` on the first `k` PCs at the corners is 0.02–0.27 for
  `k ≤ 10` and 0.96–0.97 for `k ≥ 15`; at the middle it is 0.95 already at
  `k = 10`.

So a 10-PC embedding measures "does PCA keep the causal axis" rather than
"does the method work", and CoSPAR inherits the same loss. A 20-PC embedding
(well under CYFER's identifiability cap of about 60 features at 100 clones and
3 folds) keeps the axis everywhere, and every conclusion above is drawn from
that pass.

# Problems found in the first passes, and how each was resolved

The first two passes surfaced ten problems or open decisions; Kevin reviewed
them (`additional_context/responses-to-Writeup21-trailblazing_round1.txt`) and
every one is now resolved. The `_r2` pass implements the resolutions.

1. **The embedding dimension.** Ten PCs, equal to the latent dimension, do not
   recover the ten latent coordinates, because the causal coordinate is the
   weakest direction in gene space by construction (Section 7.7). *Resolved:
   the shared embedding is 20 PCs, now the default in the drivers and written
   into memo Section 6.2 with the reasoning.*
2. **The AED re-centring step was missing** from the first implementation, and
   the 10-PC Gini sweep ran at realized AED 0.53 and 0.45 at its two lowest
   levels instead of 0.4. *Resolved: the step is in `run_sweep()` and kept;
   memo Section 5 now states the update formula explicitly
   (`h² ← clamp(h² + (realized AED − target), 0, 1)`, re-bisect `s`, up to
   three times, tolerance 0.03).*
3. **The bottom Gini level is unreachable at fixed AED 0.4**: at Gini 0.2 the
   calibrated `s = 0.21` is so small that count noise dominates the embedding,
   and even `h² = 1` leaves the realized mean AED at 0.55. *Resolved: the
   Gini ladder is re-spaced to {0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9} — starting
   above the floor and denser where the comparators fall away (memo
   Section 3.3).*
4. **Seeds.** The first implementation seeded the generator's three stages with
   `seed`, `seed + 1`, `seed + 2`, so consecutive replicates shared streams,
   and the calibration draws coincided with replicates. *Resolved and kept:
   sub-streams `10 · seed + {0, 1, 2}`, calibration draws at `+50` (memo
   Section 5).*
5. **CoSPAR and extinct clones.** With CoSPAR's default
   `extend_Tmap_space = False`, t1 cells of extinct clones fall outside the
   transition map and receive a filled-in fate bias of exactly 0.5 — and these
   are precisely the lowest-fate cells, so the fill-in ranks them in the
   middle and handicaps CoSPAR at high Gini (34–40 % of t1 cells at Gini 0.9).
   *Resolved: these cells are excluded before CoSPAR runs, and the `_r2` pass
   trailblazes that setting (Section 7.1 — CoSPAR now holds near 0.7 through
   Gini 0.6 instead of dropping below 0.3, and its Gini-0.9 score rises from
   −0.13 to about 0.3); memo Sections 3.2 and 6.3 updated.*
6. **A constant CoSPAR bias** was scored as 0 rather than reported as missing.
   *Resolved: reported as a failed run, and noted here per pass — it has not
   occurred in any dataset of any pass.*
7. **Split-half reliability** as defined in the memo (Spearman over all genes)
   cannot serve as its "well below 0.9" flag, since it is 0.2–0.7 by
   construction. *Resolved: dropped from the code and the memo's Section 7
   checklist.*
8. **Tolerances disagreed** between memo Sections 4.2 (±0.03 on the realized
   Gini during the AED sweep) and 5 (±0.05 replicate variation). *Resolved:
   memo Section 4.2 now distinguishes the calibration tolerance on the mean
   (0.01) from replicate-to-replicate sampling variation (about ±0.05).*
9. **Tooling.** `devtools::install_local()` hung on the laptop; the package
   (1.0.3.0) was installed with `R CMD INSTALL` from a scratch copy, which
   Kevin approved as the standing workaround. `Rscript` parses a driver
   lazily, so a driver is never edited while it runs.
10. **CoSPAR's ceiling of about 0.7** at the easy end of both axes means the
    figure's easy end is "CYFER 0.95, others 0.7–0.8" rather than "all three
    near 1"; Section 7.6 shows this is a property of the barcode link, not of
    the drift. *Accepted as a real finding: CoSPAR is high (around 0.7) but
    not as high as CYFER at the easy end, which is the honest picture.*

# What was and was not achieved

**Achieved.**

- The full pipeline runs end to end on the laptop across three passes (84
  datasets, plus the permutation and `τ_δ` checks), with every CYFER fit
  converged and every CoSPAR run successful.
- Both panels show the intended gradient under the settled design, with CYFER
  best at every level and the comparators falling away (CoSPAR first, then
  lineage-DE) along both axes.
- Every design decision is now settled and written into the memo, which is
  self-consistent with the trailblazing values: 20 PCs, `"noncausal"`,
  `τ_δ = 0.6`, the re-spaced Gini ladder, the AED re-centring formula, the
  seed scheme, and the CoSPAR extinct-clone exclusion (trailblazed in the
  `_r2` pass).
- The fixed values are readable from the curves: **AED 0.4** (the held value)
  is confirmed as the last easy level of its axis, and on the Gini axis the
  last easy level reads **0.6** under the extinct-clone exclusion — one notch
  above the held 0.5, a difference within two-replicate noise, so 0.5 stands
  unless the final round says otherwise (Section 7.1).
- The memo's Section 7 sanity checks are all in place, including the
  label-permutation check (every method's metric near 0 at the middle
  levels, Section 7.5).
- UMAPs of six example datasets (Section 4) give a visual sense of the data
  at the easy, middle and hard end of each axis.

**Not achieved, or still open.**

- Two replicates per level leave the top of the Gini axis noisy; the final
  round's 20 replicates are needed before anything firm is said about the top
  level, including whether the operational truth itself is the binding
  constraint there (memo Section 9).
- The final round has not run: it needs a `COSPAR_ENV` built on the Biostat
  cluster and a SLURM wrapper in the `sim3_heritability.slurm` pattern; the
  `τ_δ` supplementary row runs then.

# Anything else

- **Where the numbers live.** `csv/kevin/Writeup21/sim_{gini,aed}_{summary,
  replicate_details,calibration}{,_pca20,_r2}.csv`,
  `check_tau_delta_easy_end_pca20.csv`, `check_permutation_r2.csv`,
  `umap_coords_r2.csv`; figures in `fig/kevin/Writeup21/` (the `_r2` PNGs are
  the current ones); RDS files with the per-gene vectors of every dataset
  (`gene_stats`) in `OUT_ROOT/Writeup21_new-simulations/`.
- **Final-round settings, now settled**: 20-PC shared embedding; fixed values
  Gini 0.5 and AED 0.4; Gini ladder {0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9}; AED
  ladder as is; CoSPAR extinct-clone exclusion; 20 replicates on Hyak (which
  needs a `COSPAR_ENV` there); the `τ_δ` supplementary row at the middle
  levels.
- **A word on the truth.** Because the truth is the correlation of each gene
  with `Z_true` on the very cells the methods see, a method whose fate score
  tracks `Z_true` reproduces even the sampling noise of the 1,900 null genes;
  that is why CYFER can score above 0.9 against a truth most of whose entries
  are that sampling noise. The metric is a fair comparison between methods on
  the same cells, not an estimate of how well the genes generalize.

# References

- Chen, X. E., Lin, K. Z., Schaff, D., et al. *Resolution of Selection Versus
  Adaptation in Cellular Evolution.* Manuscript in preparation for Nature
  Methods.
- Wang, S.-W., Herriges, M. J., Hurley, K., Kotton, D. N., Klein, A. M. (2022).
  CoSpar identifies early cell fate biases from single-cell transcriptomic and
  lineage information. *Nature Biotechnology* 40, 1066–1074.
  https://doi.org/10.1038/s41587-022-01209-1
- `simulation_design_claude.md` in this directory: the design memo the round
  implements.
