---
title: "Writeup21: trailblazing round of the Gini and heterogeneity sweeps"
author: "Kevin Z. Lin (drafted by Claude)"
date: "2026-09-22"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(ggplot2)
library(knitr)
```

```{r paths}
repo_dir <- file.path("/Users/kevinlin/Library/CloudStorage/Dropbox",
                      "Collaboration-and-People/archive/Nancy/multiomeFate",
                      "git/multiomeFate_analysis")
csv_dir <- file.path(repo_dir, "csv", "kevin", "Writeup21")
```

```{r read_csvs}
.read_pass <- function(axis, suffix){
  list(summary_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_summary", suffix, ".csv")), stringsAsFactors = FALSE),
       detail_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_replicate_details", suffix, ".csv")),
    stringsAsFactors = FALSE),
       calibration_df = utils::read.csv(file.path(csv_dir,
    paste0("sim_", axis, "_calibration", suffix, ".csv")),
    stringsAsFactors = FALSE))
}
pass_list <- list(gini_10 = .read_pass("gini", ""),
                  gini_20 = .read_pass("gini", "_pca20"),
                  aed_10 = .read_pass("aed", ""),
                  aed_20 = .read_pass("aed", "_pca20"))
tau_delta_df <- utils::read.csv(file.path(csv_dir,
                                          "check_tau_delta_easy_end_pca20.csv"),
                                stringsAsFactors = FALSE)
demo_df <- utils::read.csv(file.path(csv_dir, "demo_aed_range_claude.csv"),
                           stringsAsFactors = FALSE)

method_level_vec <- c("CYFER", "CoSPAR", "Lineage-DE")
method_color_vec <- c(CYFER = "#D55E00", CoSPAR = "#0072B2",
                      "Lineage-DE" = "#009E73")
```

# Overview

This report summarizes the **trailblazing round** of the two simulation sweeps
designed in `simulation_design_claude.md` (the design memo) for Figure 4 of the
Nature Methods submission of *Resolution of Selection Versus Adaptation in
Cellular Evolution*. The round was implemented and run on 2026-09-22 on the
laptop, in two passes: the memo as written (a 10-PC shared embedding) and a
20-PC variant that the first pass showed to be necessary. Every number below is
read from the CSVs in `csv/kevin/Writeup21/`; the figures in
`fig/kevin/Writeup21/` are re-drawn here from the same tables.

| Stage | What it does | Where |
|---|---|---|
| Generate | Hierarchical latent states, one causal coordinate, negative-binomial counts for t1 cells and their t2 children | `func_generate_claude.R` |
| Calibrate | `h²` from the AED target, `s` bisected for the Gini target, `h²` re-centred on the realized AED | `calibrate_scale()`, `run_sweep()` |
| Embed | PCA on log-normalized t1 counts, t2 projected | `compute_embedding()` |
| Fit | CYFER, lineage-DE, CoSPAR, each to a per-gene association vector | `func_methods_claude.R` |
| Score | Spearman of each vector with the operational truth over 2,000 genes | `score_gene_stats()` |
| Sweep | 7 levels × 2 replicates per axis, progress file, RDS per level | `sim_gini_claude.R`, `sim_aed_claude.R` |
| Report | CSVs and barplots | `make_csvs_claude.R`, `plot_barplots_claude.R` |

# Goal

Figure 4 of the paper is planned as two panels, each a ladder of seven simulated
settings, comparing three ways of finding the genes associated with clonal
expansion from a barcoded two-time-point experiment:

- **4A, varying clone-size skewness.** As the Gini index of the later-time-point
  (t2) clone sizes rises from "many clones expanded" to "one or two clones hold
  everything", how well does each method recover the expansion genes?
- **4B, varying intraclonal heterogeneity.** As the within-clone average squared
  Euclidean distance (AED) in the embedding rises from "every cell in a clone
  looks alike" to "clone membership says nothing about a cell's state", the same
  question.

The story both panels must tell is a **gradient**: at the easy end of each axis
all three methods do well; at the hard end only CYFER should, because it models
the expansion potential of every cell rather than of clones (lineage-DE) or of
smooth state transitions tied to clone-level barcode links (CoSPAR). The
trailblazing round has a narrower goal than the figure itself: to run both
sweeps end to end at provisional fixed values (t2 Gini 0.5 held during the AED
sweep, mean squared AED 0.4 held during the Gini sweep, clone-specific t2 drift
`τ_δ = 0.6`) with two replicates per level, and to read off (i) whether the
pipeline works, (ii) where along each axis the comparators fall away, so the
final round can set its fixed values and re-space its levels, and (iii)
whether the two provisional design settings (`spread_variation = "noncausal"`
and `τ_δ = 0.6`) pan out.

# How the data are generated

One generator serves both sweeps (memo Section 2). Sizes: 100 clones, 10 t1
cells per clone, 3,000 expected t2 cells with no cap on clone size and extinct
clones kept, 2,000 genes, latent dimension `d = 10`.

**Latent state at t1.** Each clone `l` has a centre `m_l ~ N(0, τ² I)` and each
of its cells a deviation with per-coordinate standard deviation `σ_lk`. The
within-clone spread varies across clones through a multiplier
`g_l ~ Gamma(κ, κ)` with `κ = 1.5`, applied to the nine **non-causal**
coordinates only (`spread_variation = "noncausal"`), so a clone's spread on the
expansion axis, and hence its expected t2 size, is the same for tight and
diffuse clones. The two knobs are a scale and a share: `τ = s · sqrt(h²)`,
`σ_w = s · sqrt(1 − h²)`.

**Fate potential.** The first coordinate is the expansion axis:
`Z_i = β_0 + β s_i1` with `β = 1.5`, `β_0` solved so the expected number of t2
cells is 3,000, and each t1 cell leaves `Poisson(exp(Z_i))` children. The t2
size of a clone is the sum over its cells, zeros included.

**Genes.** 100 expansion genes load on the causal coordinate (magnitudes
`Unif(0.5, 1)`, half positive, half negative); every gene loads on the other
nine coordinates with dense `N(0, 0.25²)` loadings. Counts are negative
binomial (`θ = 10`) with log-normal library sizes around 5,000 and `N(0, 1)`
log-baselines.

**The later time point.** The t2 population is the multiset of children. A
child's latent state is `ρ · s_parent + (1 − ρ) · m_l + sqrt(1 − ρ²) σ_l ∘ e +
δ + δ_l` with `ρ = 0.8`, a shared shift `δ` of norm `3s` on the non-causal
coordinates, and a clone-specific shift `δ_l ~ N(0, τ_δ² I)` with `τ_δ = 0.6`.
Only CoSPAR ever sees t2 expression.

**Seeds.** Replicate `r` of level `j` uses seed `seed_base + 100 j + r`
(`seed_base` 1000 for the Gini axis, 2000 for the AED axis); the generator's
three stochastic stages use `10 · seed + {0, 1, 2}`; the calibration draws use
`seed_base + 100 j + 50 + k`. (The first version of the code reused the
calibration draws as replicates and shared sub-streams between consecutive
replicates; see Section 6.)

# How the simulation suite is set up

## The two axes and their calibration

Both axes are stated in terms of a **realized** statistic, not a generator
parameter:

- **Gini axis:** the t2-only Gini of the 100 clone sizes, zeros included
  (`gini_coef()`, identical to `dineq::gini.wtd`), targets
  `{0.2, 0.3, 0.4, 0.5, 0.6, 0.75, 0.9}`, mean squared AED held at 0.4.
- **AED axis:** the mean over clones of the per-clone squared AED
  (`SqDist_l / SqDist_random` in the shared embedding, computed as twice the
  sum of per-coordinate variances), targets
  `{0.1, 0.25, 0.4, 0.55, 0.7, 0.85, 1.0}`, t2 Gini held at 0.5.

Calibration per level: `h² = 1 − AED target`; `s` bisected (geometric, mean
Gini over 10 latent-only draws, tolerance 0.01); then one full dataset is
generated and embedded, its realized mean AED compared with the target, and
`h²` shifted by the miss up to three times (tolerance 0.03), re-bisecting `s`
each time. The re-centring step was added after the code review (Section 6);
the 10-PC pass ran without it.

```{r calibration_tables}
.calib_table <- function(df, axis_label){
  out_df <- df[, c("level", "h2_initial", "h2", "num_recentre", "aed_check",
                   "latent_scale", "tau", "sigma_w", "gini_calibrated")]
  out_df[sapply(out_df, is.numeric)] <- lapply(out_df[sapply(out_df, is.numeric)],
                                               round, 3)
  colnames(out_df) <- c(axis_label, "h2 initial", "h2 final", "re-centrings",
                        "check AED", "s", "tau", "sigma_w", "Gini (10 draws)")
  out_df
}
knitr::kable(.calib_table(pass_list$gini_20$calibration_df, "Gini target"),
             caption = "Calibration of the Gini axis (20-PC pass, with re-centring).")
knitr::kable(.calib_table(pass_list$aed_20$calibration_df, "AED target"),
             caption = "Calibration of the AED axis (20-PC pass, with re-centring).")
```

Two things the tables show. The calibrated latent scale `s` runs from 0.21 to
2.06 along the Gini axis: the Gini is moved by making the fate potential
`β s_i1` more or less spread across cells, so the top Gini level is also the
level with the strongest expression signal and the bottom level the weakest.
And the re-centring hits its wall at the bottom of both axes: the Gini-0.2
level needs `h² = 1` (no within-clone latent spread at all) and still realizes a
mean AED of 0.57 rather than 0.4, because at `s = 0.21` the count noise
dominates the embedding; the AED-0.1 level likewise needs `h² = 1` and realizes
0.13. Every other level lands within 0.03 of its target.

## Runs

Two passes, each 7 levels × 2 replicates × 2 axes = 28 datasets:

| Pass | Embedding | Calibration | Seeds | Output suffix | Role |
|---|---|---|---|---|---|
| 1 | 10 PCs (memo Section 6.2) | no AED re-centring | original scheme | none | the memo as written; the record of why the embedding dimension matters |
| 2 | 20 PCs | with re-centring | corrected scheme | `_pca20` | the pass whose curves answer the trailblazing questions |

One dataset takes about a minute (generation 1 s, embedding 5 s, CYFER 3–6 s,
lineage-DE 1 s, CoSPAR 15–45 s); a sweep is 12–13 minutes per axis and the two
axes run concurrently. The 10-PC pass also served as the dry test (2 levels × 1
replicate first, then the full grid).

# Methods compared

All three methods see the same 1,000 t1 cells, the same clone labels, the same
t2 clone sizes and, where they use one, the same embedding (PCA of
log-normalized t1 counts; t2 cells projected with the t1 centre and rotation).

**CYFER.** `cyfer()` with 3-fold cross-validation over a 10-value λ path and
`cyfer_finalize()` on the PCs, with the zero-count clones **kept** (the
sim1–sim7 scripts drop them; here they carry the signal at high Gini). The
per-gene statistic is the Spearman correlation of each gene's log-normalized
expression with `cell_imputed_score` over t1 cells.

**Lineage-DE.** Clones are split into "high" (t2 size above the *mean*) and
"low"; every gene is compared between the t1 cells of high and low clones by
the Wilcoxon rank-sum test, and the statistic is the rank-biserial correlation
`2U / (n_H n_L) − 1`, computed for all genes at once from ranks and
cross-checked against `stats::wilcox.test()` on five genes per dataset (U and
p-value agree to 1e-6).

**CoSPAR.** The matched route: `state_info` is `"t1"` for t1 cells and
`"High"` / `"Low"` for t2 cells by the same mean split;
`infer_Tmap_from_multitime_clones()` on the shared PCA (paper defaults:
`smooth 20 15 10 5`, `sparsity 0.1`, `intraclone 0.2`); `fate_bias(High, Low)`
on the intraclone map; the statistic is the Spearman correlation of each gene
with the fate bias over t1 cells. CoSPAR runs in its own conda environment
through the `run-cospar` flat-file templates.

**The metric.** The truth is operational: `truth_g = Spearman(x_g, Z_true)`
over the same t1 cells. The score of a method is
`Spearman(m, truth)` over all 2,000 genes (Pearson and a Jaccard of BH
`q < 0.05` sets are secondary columns). Per dataset the tables also record the
correlation of each method's per-cell fate score (`Z_hat`, fate bias) with
`Z_true`, so a gene-level failure can be traced to the score or to the gene
step.

# Results

## The headline curves (20-PC pass)

```{r plot_helpers}
.level_label_vec <- function(summary_df, axis){
  level_df <- unique(summary_df[, c("level_idx", "level", "gini_t2_mean",
                                    "aed_mean_mean", "aed_q05_mean",
                                    "aed_q95_mean")])
  level_df <- level_df[order(level_df$level_idx), ]
  if(axis == "gini"){
    label_vec <- paste0(level_df$level, "\n(",
                        sprintf("%.2f", level_df$gini_t2_mean), ")")
  } else {
    label_vec <- paste0(level_df$level, "\n(",
                        sprintf("%.2f", level_df$aed_mean_mean), ")\n[",
                        sprintf("%.2f", level_df$aed_q05_mean), ", ",
                        sprintf("%.2f", level_df$aed_q95_mean), "]")
  }
  names(label_vec) <- level_df$level_idx
  label_vec
}

.plot_barplot <- function(summary_df, detail_df, axis, metric, ylab, title){
  summary_df <- summary_df[summary_df$method %in% method_level_vec, ]
  detail_df <- detail_df[detail_df$method %in% method_level_vec, ]
  summary_df$method <- factor(summary_df$method, levels = method_level_vec)
  detail_df$method <- factor(detail_df$method, levels = method_level_vec)
  summary_df$level_factor <- factor(summary_df$level_idx)
  detail_df$level_factor <- factor(detail_df$level_idx)
  mean_col <- paste0(metric, "_mean")
  label_vec <- .level_label_vec(summary_df, axis)
  x_label <- if(axis == "gini"){
    "Gini of t2 clone sizes: target (realized mean)"
  } else {
    "Mean squared AED: target (realized mean) [per-clone 5th-95th pct]"
  }

  plot1 <- ggplot2::ggplot(summary_df,
                           ggplot2::aes(x = level_factor, y = .data[[mean_col]],
                                        fill = method))
  plot1 <- plot1 + ggplot2::geom_col(position = ggplot2::position_dodge(0.8),
                                     width = 0.7)
  plot1 <- plot1 + ggplot2::geom_point(
    data = detail_df,
    ggplot2::aes(x = level_factor, y = .data[[metric]], group = method),
    position = ggplot2::position_dodge(0.8), size = 1.2, color = "black",
    inherit.aes = FALSE)
  plot1 <- plot1 + ggplot2::geom_hline(yintercept = 0, linewidth = 0.3)
  plot1 <- plot1 + ggplot2::scale_fill_manual(values = method_color_vec)
  plot1 <- plot1 + ggplot2::scale_x_discrete(labels = label_vec)
  plot1 <- plot1 + ggplot2::labs(x = x_label, y = ylab, fill = NULL,
                                 title = title)
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot1
}

.summary_table <- function(summary_df, axis_label){
  wide_df <- data.frame(level = sort(unique(summary_df$level)))
  for(method in method_level_vec){
    sub_df <- summary_df[summary_df$method == method, ]
    sub_df <- sub_df[match(wide_df$level, sub_df$level), ]
    wide_df[[method]] <- sprintf("%.2f (%.2f)", sub_df$spearman_mean,
                                 sub_df$spearman_sd)
  }
  colnames(wide_df)[1] <- axis_label
  wide_df
}
```

```{r gini_20_bar, fig.width=9, fig.height=5}
plot(.plot_barplot(pass_list$gini_20$summary_df, pass_list$gini_20$detail_df,
                   "gini", "spearman", "Spearman(method, truth) over genes",
                   "4A: Gini axis, 20-PC embedding"))
```

```{r aed_20_bar, fig.width=9, fig.height=5}
plot(.plot_barplot(pass_list$aed_20$summary_df, pass_list$aed_20$detail_df,
                   "aed", "spearman", "Spearman(method, truth) over genes",
                   "4B: AED axis, 20-PC embedding"))
```

```{r summary_tables_20}
knitr::kable(.summary_table(pass_list$gini_20$summary_df, "Gini target"),
             caption = "Gini axis, 20 PCs: Spearman of each method's per-gene vector with the truth, mean (SD) over 2 replicates.")
knitr::kable(.summary_table(pass_list$aed_20$summary_df, "AED target"),
             caption = "AED axis, 20 PCs: Spearman of each method's per-gene vector with the truth, mean (SD) over 2 replicates.")
```

**Both panels show the intended gradient.** On the Gini axis CYFER stays at
0.88–0.95 from Gini 0.3 to 0.75 while CoSPAR falls from about 0.7 to 0.28 at
Gini 0.6 and to −0.13 at 0.9, and lineage-DE from about 0.75 to 0.56–0.60. On
the AED axis CYFER stays at 0.79–0.97 across the whole ladder while CoSPAR
falls from 0.7 to −0.04 and lineage-DE from 0.75–0.84 to 0.23. CYFER is the
best method at every level of both axes.

Two qualifications. At the top Gini level (0.9) CYFER's gene-level score drops
to 0.52 (SD 0.05) even though its fate score still correlates 0.82 with the
truth: one clone holds about half of the 3,000 t2 cells and 25 clones are
extinct, so the operational truth over 1,000 cells is itself dominated by a few
dozen cells. And at the bottom Gini level (0.2) all three methods sit at
0.67–0.75: this is the weak-signal regime (`s = 0.21`), where the fate
potential barely varies and the realized AED could not be brought below 0.55
(Section 6).

## The fate scores behind the gene calls

```{r score_bars, fig.width=9, fig.height=4.5}
plot(.plot_barplot(pass_list$gini_20$summary_df, pass_list$gini_20$detail_df,
                   "gini", "score_cor",
                   "cor(fate score, true fate potential) over t1 cells",
                   "Gini axis, 20 PCs: per-cell fate scores"))
plot(.plot_barplot(pass_list$aed_20$summary_df, pass_list$aed_20$detail_df,
                   "aed", "score_cor",
                   "cor(fate score, true fate potential) over t1 cells",
                   "AED axis, 20 PCs: per-cell fate scores"))
```

CYFER's `cell_imputed_score` correlates 0.74–0.98 with `Z_true` at every level
of both axes (0.82 at Gini 0.9). CoSPAR's fate bias tops out near 0.7 at the
easy end of both axes and decays to about 0.1 at the hard end; its ceiling is
not the t2 drift (next subsection). Lineage-DE has no per-cell score.

## Secondary metrics

```{r secondary_tables}
.secondary_table <- function(summary_df, axis_label){
  out_df <- summary_df[order(summary_df$level, summary_df$method),
                       c("level", "method", "pearson_mean", "jaccard_mean",
                         "num_converged")]
  out_df$pearson_mean <- round(out_df$pearson_mean, 2)
  out_df$jaccard_mean <- round(out_df$jaccard_mean, 2)
  colnames(out_df) <- c(axis_label, "method", "Pearson", "Jaccard (q < 0.05)",
                        "converged / 2")
  out_df
}
knitr::kable(.secondary_table(pass_list$gini_20$summary_df, "Gini target"),
             caption = "Gini axis, 20 PCs: secondary metrics, mean over 2 replicates.")
knitr::kable(.secondary_table(pass_list$aed_20$summary_df, "AED target"),
             caption = "AED axis, 20 PCs: secondary metrics, mean over 2 replicates.")
```

Every CYFER fit and every CoSPAR run converged (56 datasets across the two
passes, no `NULL` fit and no collapsed bias). The Jaccard column orders the
methods the same way as the Spearman headline at almost every level but is
noisier, as expected from a thresholded set.

## Realized statistics and diagnostics

```{r realized_table}
.realized_table <- function(summary_df, axis_label){
  out_df <- unique(summary_df[, c("level", "gini_t2_mean", "aed_mean_mean",
                                  "aed_q05_mean", "aed_q95_mean",
                                  "t2_max_clone_mean", "t2_num_zero_mean",
                                  "num_high_clones_mean",
                                  "spearman_aed_t2size_mean",
                                  "split_half_pearson_mean",
                                  "pca_heritability_mean")])
  out_df <- out_df[order(out_df$level), ]
  out_df[sapply(out_df, is.numeric)] <- lapply(out_df[sapply(out_df, is.numeric)],
                                               round, 2)
  colnames(out_df) <- c(axis_label, "Gini t2", "mean AED", "AED 5th pct",
                        "AED 95th pct", "largest clone", "extinct clones",
                        "high clones", "Spearman(AED_l, Y_l)",
                        "split-half (Pearson)", "PCA heritability")
  out_df
}
knitr::kable(.realized_table(pass_list$gini_20$summary_df, "Gini target"),
             caption = "Gini axis, 20 PCs: realized statistics per level, mean over 2 replicates.")
knitr::kable(.realized_table(pass_list$aed_20$summary_df, "AED target"),
             caption = "AED axis, 20 PCs: realized statistics per level, mean over 2 replicates.")
```

- **The calibration holds** except at the bottom of the Gini axis: the realized
  Gini is within 0.03 of its target at every level of both axes (and within
  0.06 of the fixed 0.5 during the AED sweep), and the realized mean AED is
  within 0.03 of its target everywhere except Gini 0.2 (0.55 against 0.4).
- **The per-clone AED spans the intended range**: at the top AED level the
  5th–95th percentiles are 0.27–1.71 (the per-clone maximum exceeds 2).
- **The spread-size coupling is gone**, as the `"noncausal"` placement was meant
  to achieve: Spearman between a clone's AED and its t2 size is between −0.10
  and 0.11 at every level of both axes.
- **The Gini axis reaches the intended regime**: at Gini 0.9 the largest clone
  holds about 1,400 of 3,000 t2 cells and 25 clones are extinct; the lineage-DE
  high group shrinks from 47 clones to 14.
- **Split-half reliability.** The memo's Spearman version of the truth's
  split-half reliability is 0.2–0.7 by construction (1,900 of the 2,000 genes
  have a null truth, so the rank correlation between halves is mostly noise);
  the Pearson version, which weights the ~100 signal genes, is 0.52–0.91 and
  rises with `s`. Neither flags a level as unrecoverable.

## Does the clone-specific t2 drift explain CoSPAR's ceiling?

Memo Section 9 says that if CoSPAR does poorly even at the easy end, `τ_δ` is
the first thing to lower. At the easiest AED level (`h² = 0.9`, Gini 0.5, two
seeds, 20 PCs) CoSPAR was re-run at `τ_δ ∈ {0, 0.3, 0.6, 1.2}`:

```{r tau_delta_table}
td_df <- tau_delta_df[tau_delta_df$method == "CoSPAR", ]
td_wide_df <- data.frame(tau_delta = sort(unique(td_df$tau_delta)))
for(seed in sort(unique(td_df$seed))){
  sub_df <- td_df[td_df$seed == seed, ]
  sub_df <- sub_df[match(td_wide_df$tau_delta, sub_df$tau_delta), ]
  td_wide_df[[paste0("Spearman, seed ", seed)]] <- round(sub_df$spearman, 2)
  td_wide_df[[paste0("cor(bias, Z_true), seed ", seed)]] <- round(sub_df$score_cor, 2)
}
knitr::kable(td_wide_df,
             caption = "CoSPAR at the easiest AED level as the clone-specific t2 drift varies.")
```

CoSPAR is essentially insensitive to `τ_δ`: its fate-score correlation moves
from 0.72 to 0.65–0.67 over a four-fold range of drift, and its gene-level score
by about the same. Its ceiling of roughly 0.7 on these data is therefore the
uniform barcode link (every t1 cell of a clone tied equally to every t2 cell of
that clone), not the loss of cross-clone coherence at t2. `τ_δ = 0.6` can
stand, and the paper's framing should say "uniform link and extinct clones"
rather than "smooth continuum" (memo Section 9, second bullet, second branch).

## The 10-PC pass: why the embedding dimension matters

```{r compare_passes, fig.width=9, fig.height=4.5}
.compare_df <- function(axis){
  a_df <- pass_list[[paste0(axis, "_10")]]$summary_df
  b_df <- pass_list[[paste0(axis, "_20")]]$summary_df
  a_df$pass <- "10 PCs"
  b_df$pass <- "20 PCs"
  out_df <- rbind(a_df[, c("level", "method", "pass", "spearman_mean",
                           "score_cor_mean")],
                  b_df[, c("level", "method", "pass", "spearman_mean",
                           "score_cor_mean")])
  out_df$method <- factor(out_df$method, levels = method_level_vec)
  out_df
}
for(axis in c("gini", "aed")){
  cmp_df <- .compare_df(axis)
  plot1 <- ggplot2::ggplot(cmp_df, ggplot2::aes(x = level, y = spearman_mean,
                                                color = method,
                                                linetype = pass))
  plot1 <- plot1 + ggplot2::geom_line()
  plot1 <- plot1 + ggplot2::geom_point()
  plot1 <- plot1 + ggplot2::scale_color_manual(values = method_color_vec)
  plot1 <- plot1 + ggplot2::labs(
    x = if(axis == "gini") "Gini target" else "Mean squared AED target",
    y = "Spearman(method, truth), mean of 2", color = NULL, linetype = NULL,
    title = paste0(if(axis == "gini") "Gini" else "AED",
                   " axis: 10-PC pass (memo as written) versus 20-PC pass"))
  plot1 <- plot1 + ggplot2::theme_bw()
  plot1 <- plot1 + ggplot2::theme(legend.position = "top")
  plot(plot1)
}
```

```{r compare_table}
cmp_list <- lapply(c("gini", "aed"), function(axis){
  cmp_df <- .compare_df(axis)
  cmp_df <- cmp_df[cmp_df$method == "CYFER", ]
  wide_df <- data.frame(axis = axis, level = sort(unique(cmp_df$level)))
  for(pass in c("10 PCs", "20 PCs")){
    sub_df <- cmp_df[cmp_df$pass == pass, ]
    sub_df <- sub_df[match(wide_df$level, sub_df$level), ]
    wide_df[[paste0("genes, ", pass)]] <- round(sub_df$spearman_mean, 2)
    wide_df[[paste0("fate score, ", pass)]] <- round(sub_df$score_cor_mean, 2)
  }
  wide_df
})
knitr::kable(do.call(rbind, cmp_list),
             caption = "CYFER under the two embeddings: gene-level Spearman and fate-score correlation. (The 10-PC pass predates the re-centring and seed fixes, so its levels are not identical datasets.)")
```

With the memo's 10-PC embedding CYFER collapses at the hard end of both axes
(gene-level 0.06 at Gini 0.9 and 0.09 at AED 1.0, fate-score correlation 0.06
and 0.01) and at the weak-signal bottom of the Gini axis (0.44), and
lineage-DE, which never uses the embedding, looks best at the top. The
diagnosis, established with side experiments during the run:

- Fitting CYFER on the **true latent coordinates** instead of the PCs gives a
  fate-score correlation of 1.00 at every setting, so the model is not the
  problem.
- The causal axis loads on 100 genes with squared loadings around 0.58, while
  each of the nine non-causal coordinates loads on all 2,000 genes with squared
  loadings 0.0625, so in gene space the causal axis carries about half the
  variance of any non-causal one and is always the **weakest latent
  direction**. At the middle of either axis it sits on PC 10 (correlation
  0.94 with the causal coordinate); at the corners it drops to PC 11
  (correlation 0.83–0.96 with PC 11, at most 0.31 with any of PCs 1–10).
- The R² of `Z_true` on the first `k` PCs at the corners is 0.02–0.27 for
  `k ≤ 10` and 0.96–0.97 for `k ≥ 15`; at the middle it is 0.95 already at
  `k = 10`.

So a 10-PC embedding measures "does PCA keep the causal axis" rather than
"does the method work", and CoSPAR inherits the same loss. A 20-PC embedding
(well under CYFER's identifiability cap of about 60 features at 100 clones and
3 folds) keeps the axis everywhere, and every conclusion above is drawn from
that pass.

# Problems and potential issues found while running

1. **The embedding dimension** (above). The memo fixes 10 PCs everywhere, equal
   to the latent dimension, on the assumption that 10 PCs recover 10 latent
   coordinates; they do not, because the causal coordinate is the weakest
   direction in gene space by construction. This is the one design change the
   memo needs. Alternatives to 20 PCs would be to shrink the non-causal
   loadings or to spread the causal loading over more genes, either of which
   changes the gene model.
2. **The AED re-centring step was missing** from the first implementation
   (memo Section 5 asks for it), and the 10-PC Gini sweep ran at realized AED
   0.53 and 0.45 at its two lowest levels instead of 0.4. The step is now in
   `run_sweep()`.
3. **The bottom Gini level is unreachable at fixed AED 0.4.** At Gini 0.2 the
   calibrated `s = 0.21` is so small that count noise dominates the embedding,
   and even `h² = 1` (no within-clone latent spread) leaves the realized mean
   AED at 0.55. The Gini ladder should start at 0.3 (which needs `h² = 0.785`).
   Symmetrically, AED 0.1 needs `h² = 1` at 20 PCs.
4. **Seeds.** The first implementation seeded the generator's three stages with
   `seed`, `seed + 1`, `seed + 2`, so consecutive replicates shared streams,
   and the calibration draws at `seed + k` coincided with replicates
   `k ≤ 10`. Fixed (`10 · seed + {0, 1, 2}`, calibration at `+50`); the 20-PC
   pass was re-run with the fix.
5. **CoSPAR and extinct clones.** `run_cospar.py` uses CoSPAR's default
   `extend_Tmap_space = False`, so t1 cells of clones with no t2 cells fall
   outside the map and receive a fate bias of exactly 0.5 (34–40 % of t1 cells
   at Gini 0.9), whereas memo Section 3.2 says their bias "comes entirely from
   smoothing over neighbours". This is a fairness decision, not a code bug:
   either the memo's sentence changes or the flag is set. Note that these
   cells are the lowest-fate cells, and a bias of 0.5 ranks them in the middle,
   which handicaps CoSPAR at high Gini.
6. **A constant CoSPAR bias** was scored as 0 rather than reported as missing;
   now reported as a failed run (it did not occur in any dataset).
7. **Split-half reliability** as defined in the memo (Spearman over all genes)
   cannot serve as its "well below 0.9" flag, since it is 0.2–0.7 by
   construction; the Pearson version is recorded beside it.
8. **Tolerances disagree** between memo Sections 4.2 (realized Gini within
   ±0.03 of the fixed value during the AED sweep) and 5 (±0.05 replicate
   variation); the AED-sweep replicates run 0.45–0.55.
9. **Tooling.** `devtools::install_local()` hung on the laptop; the package
   (1.0.3.0) was installed with `R CMD INSTALL` from a scratch copy. `Rscript`
   parses a driver lazily, so editing a driver while it runs breaks the run at
   the edited point (it happened once, after the RDS was saved).
10. **CoSPAR's ceiling of about 0.7** at the easy end of both axes means the
    figure's easy end is "CYFER 0.95, others 0.7–0.8" rather than "all three
    near 1"; Section 5.5 shows this is a property of the barcode link, not of
    the drift, so it is real rather than an artefact of `τ_δ`.

# What was and was not achieved

**Achieved.**

- The full pipeline runs end to end on the laptop: generator, calibration,
  three methods, scoring, progress file, RDS, CSVs, figures; 56 datasets with
  no failed fit.
- Both panels show the intended gradient under the 20-PC embedding, with CYFER
  best at every level and the comparators falling away (CoSPAR first, then
  lineage-DE) along both axes.
- The provisional design settings pan out: `"noncausal"` removes the
  spread-size coupling without narrowing the per-clone AED range, and
  `τ_δ = 0.6` is not what limits CoSPAR.
- The "last easy level" the final round needs is readable: on the corrected
  curves it is **Gini 0.5** (all three ≥ 0.77; CoSPAR drops to 0.28 at 0.6) and
  **AED 0.4** (all three ≥ 0.70; CoSPAR 0.47 and lineage-DE 0.51 at 0.55), which
  are the memo's provisional fixed values.
- The realized-statistic bookkeeping the memo asks for (Gini, AED and its
  quantiles, extinct clones, largest clone, coupling, heritability, group
  sizes, convergence) is in every replicate row.

**Not achieved, or changed.**

- The memo's 10-PC embedding does not produce a usable figure; the result
  stands only under a 20-PC embedding, which the memo has to adopt (or the gene
  model has to change).
- The Gini-0.2 level cannot be run at the fixed AED, so the ladder is six usable
  levels until it is re-spaced.
- The memo's Section 7 sanity checks were not run as a block by instruction;
  the label-permutation check at the middle level is the one item with no
  corresponding column.
- Two replicates per level leave the top of the Gini axis noisy (lineage-DE's
  SD is 0.35 at Gini 0.9, CoSPAR's 0.58); the final round's 20 replicates are
  needed before anything is said about the top level.
- The CoSPAR extinct-clone convention is unresolved.

# Anything else

- **Where the numbers live.** `csv/kevin/Writeup21/sim_{gini,aed}_{summary,
  replicate_details,calibration}{,_pca20}.csv`,
  `check_tau_delta_easy_end_pca20.csv`; figures in `fig/kevin/Writeup21/`; RDS
  files with the per-gene vectors of every dataset (`gene_stats`) in
  `OUT_ROOT/Writeup21_new-simulations/`.
- **Suggested final-round settings**, pending Kevin's decisions: 20-PC shared
  embedding; fixed values Gini 0.5 and AED 0.4; Gini ladder starting at 0.3 with
  denser spacing between 0.5 and 0.9; AED ladder as is; 20 replicates on Hyak
  (which needs a `COSPAR_ENV` there); the `τ_δ` supplementary row at the
  middle levels.
- **A word on the truth.** Because the truth is the correlation of each gene
  with `Z_true` on the very cells the methods see, a method whose fate score
  tracks `Z_true` reproduces even the sampling noise of the 1,900 null genes;
  that is why CYFER can score 0.95 against a truth whose Spearman split-half
  reliability is 0.6. The metric is a fair comparison between methods on the
  same cells, not an estimate of how well the genes generalize.

# References

- Chen, X. E., Lin, K. Z., Schaff, D., et al. *Resolution of Selection Versus
  Adaptation in Cellular Evolution.* Manuscript in preparation for Nature
  Methods.
- Wang, S.-W., Herriges, M. J., Hurley, K., Kotton, D. N., Klein, A. M. (2022).
  CoSpar identifies early cell fate biases from single-cell transcriptomic and
  lineage information. *Nature Biotechnology* 40, 1066–1074.
  https://doi.org/10.1038/s41587-022-01209-1
- `simulation_design_claude.md` in this directory: the design memo the round
  implements.
